### 2022-12-03 manuscript/manuscript.Rmd
updated the figure and some references.

---
title: "Expanding the scope of experimental archaeology using the Perception-Process-Product analytical framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: "This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology. \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Collaborative knowledge production"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

# Introduction

This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology (**Figure** \@ref(fig:concept)). Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016]. In the past decades, actualistical experiments becomes more common [@liu2022].

```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="The conceptual diagram of the Perception-Process-Product analytical framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```

Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidle2009; @haidle2010; @lombard2012], essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal. This approach is innovative yet limited by the curse of expertise [@hinds1999]. Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation. Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014].

# Many places, many voices

The PPP analytical framework inherently advocate an collaborative mode of knowledge production [@ranhorn2020].

# References

### 2022-12-05 manuscript/manuscript.Rmd
updated the outline of this article.

@@
 This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology (**Figure** \@ref(fig:concept)).
+minimal engineering using raw materials available in the past to demonstrate it is possible to do something.
+
+This part talks about the goal and toolbox of PPP framework, which is understanding the multi-level understanding of variation. the first two p captures different level of variation
+
 Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016].
@@
 
-Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidle2009; @haidle2010; @lombard2012], essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
-This approach is innovative yet limited by the curse of expertise [@hinds1999].
+Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidle2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
+This approach is a power and elegant yet limited by the curse of expertise [@hinds1999].
+Like chaine operatoire, it cannot handles variation very well.
+To some extent it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps.
 Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation.
@@
 
+# The curse of knowledge
+
+Variation: why novice is important?
+
 # Many places, many voices
+
+Variation: why experts or collaborative knowledge from different regions matters?
 

### 2022-12-06 manuscript/manuscript.Rmd
added a few more references.

@@
 
-This part talks about the goal and toolbox of PPP framework, which is understanding the multi-level understanding of variation. the first two p captures different level of variation
+This part talks about the goal and toolbox of PPP framework, which is understanding the multi-level understanding of variation.
+the first two p captures different level of variation: EQUIFINALITY [@chami2015].
+
+Is RCT the golden standard of knowledge [@cartwright2007]
 
@@
 In the past decades, actualistical experiments becomes more common [@liu2022].
+new toolkit such as BORIS were introduced [@friard2016]
 
@@
 
-The PPP analytical framework inherently advocate an collaborative mode of knowledge production [@ranhorn2020].
+The PPP analytical framework inherently adopts an collaborative mode of knowledge production, which has been advocated in experimental studies [@ranhorn2020] and museum collection studies [@timbrell2022].
+
+# Open science beyond reproducibility
+
+The last step is uploading the data to a open-access repository [@marwick2017]. The building of manufacture can cost.
 

### 2023-03-07 manuscript/manuscript.Rmd
add some references.
adding some texts.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
-abstract: "This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology.
+abstract: "This paper presents the Perception-Process-Product analytical framework that aims to expand the scope of experimental archaeology .
 \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Collaborative knowledge production"
@@
 
-This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology (**Figure** \@ref(fig:concept)).
+This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology that mainly focuses on the product level (**Figure** \@ref(fig:concept)).
+It aims to expose the behavioral variability at multiple levels and build the connections.
 minimal engineering using raw materials available in the past to demonstrate it is possible to do something.
@@
 the first two p captures different level of variation: EQUIFINALITY [@chami2015].
+The process level can be dismantled into two parts: 1)the cognitive and physical demands of certain tool-making behavior; 2) the ethological analysis.
 
@@
 
-Variation: why novice is important?
+Variation: why novice is important? [@hinds1999]
 
@@
 
-The last step is uploading the data to a open-access repository [@marwick2017]. The building of manufacture can cost.
+The last step is uploading the data to a open-access repository [@marwick2017]. The building of manufacture can cost [@simon2015; @gilmore2015].
 

### 2023-03-08 manuscript/manuscript.Rmd
outline finished.

--- 2022-12-06 manuscript/manuscript.Rmd
@@
 ---
-title: "Expanding the scope of experimental archaeology using the Perception-Process-Product analytical framework"
+title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
 author: 
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
-abstract: "This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology.
-\\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Collaborative knowledge production"
+abstract: "This paper presents the Perception-Process-Product analytical framework that aims to expand the scope of experimental archaeology .
+\\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
 date: "`r Sys.Date()`"
@@
 
-This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology (**Figure** \@ref(fig:concept)).
-minimal engineering using raw materials available in the past to demonstrate it is possible to do something.
+This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to centers around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+The Triple P framework aims to amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)).
+To accomplish these two objectives, we advocate the following four measures: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse.
+It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspires researchers to ask questions that are bold and transformative [@schmidt2020].
+Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.
 
-This part talks about the goal and toolbox of PPP framework, which is understanding the multi-level understanding of variation.
-the first two p captures different level of variation: EQUIFINALITY [@chami2015].
-
+Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020].
+In the past decades, actualistic experiments becomes more common [@liu2022].
 Is RCT the golden standard of knowledge [@cartwright2007]
 
-Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016].
-In the past decades, actualistical experiments becomes more common [@liu2022].
+the first two p captures different level of variation: EQUIFINALITY [@chami2015].
+The process level can be dismantled into two parts: 1)the cognitive [@stout2015] and physical demands [@key2018] of certain tool-making behavior; 2) the ethological analysis.
+
+# The ethology and ethnography of stone toolmaking
+
 new toolkit such as BORIS were introduced [@friard2016]
@@
 
-Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidle2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
+Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
 This approach is a power and elegant yet limited by the curse of expertise [@hinds1999].
@@
 Variation: why novice is important?
+[@hinds1999].
+Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979].
 

### 2023-03-10 manuscript/manuscript.Rmd
added some text.

@@
 
+Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping [@whittaker2004] are not novel, however, it has never been formally recognized as a legitimate research method in experimental archaeology.
+
 # The curse of knowledge
 
-Variation: why novice is important?
-[@hinds1999].
+We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with strong academic background, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
+When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits, preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
+The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
+However, what is problematic is that the results of replication experiments conducted by these experienced practioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
+
 Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979].
@@
 
-Variation: why experts or collaborative knowledge from different regions matters?
-
-The PPP analytical framework inherently adopts an collaborative mode of knowledge production, which has been advocated in experimental studies [@ranhorn2020] and museum collection studies [@timbrell2022].
+Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020] and museum collection studies [@timbrell2022] of stone artifacts.
 

### 2023-03-11 manuscript/manuscript.Rmd
added some text.

@@
 
+One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition.
+The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity.
+Roe and Just (2009: 1266-1267) defined internal validity as “the ability of a researcher to argue that observed correlations are causal” and external validity as “the ability to generalize the relationships found in a study to other persons, times, and settings”.
+This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs.
+In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades (Eren et al., 2016).
+Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area.
+All the variables of interest are easy to measure, quantify, and control in a machine knapping setting.
+Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions.
+In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in teaching experiments.
+As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.
+
+Experimental archaeology is based on the concept of analogy (i.e., the past is at least partially similar to the present in some aspects).
+It is acknowledged that the validity of this type of analogical inference has long been a subject of debate in archaeology (Wylie, 1985), and a comprehensive review of it is beyond the scope of this essay.
+For instance, presumably in the paleolithic period the development of stone knapping skills mostly happened during childhood and these children grew up in an environment surrounded by habitual stone tool users and makers.
+In this case, what can knapping teaching experiments involving modern adults who have zero exposure to stone tools inform us about the past learning behavior?
+It is important to clarify that no experimental project ever intends or claims to provide the perfect reconstruction of the past but rather aims at identifying variables relevant to the question of interest (Stout & Khreisheh, 2015), which can be often ignored through pure deductive reasoning.
+In the end, all experiments are wrong, but some are useful, and we need more of them.
+
 # The ethology and ethnography of stone toolmaking

### 2023-03-14 manuscript/manuscript.Rmd
minor revision.

@@
 
-This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to centers around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-The Triple P framework aims to amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)).
+This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)).
 To accomplish these two objectives, we advocate the following four measures: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse.
@@
 The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity.
-Roe and Just (2009: 1266-1267) defined internal validity as “the ability of a researcher to argue that observed correlations are causal” and external validity as “the ability to generalize the relationships found in a study to other persons, times, and settings”.
+Roe and Just (2009: 1266-1267) defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings".
 This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs.
@@
 
-```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="The conceptual diagram of the Perception-Process-Product analytical framework."}
+```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram of the Perception-Process-Product conceptual framework."}
 knitr::include_graphics(here::here("figure/Fig1.jpg"))

### 2023-03-15 manuscript/manuscript.Rmd
added some text.

@@
 
-We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with strong academic background, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
@@
 
+In addition to the difficulty in coordination and logistics, the facilitation of large-scale collaborations is often hindered by the current system of research evaluation, where usually only the first author and the senior (last/correspondent) author of a peer-reviewed journal paper will be acknowledged as proper contribution.
+
 # Open science beyond reproducibility

### 2023-03-17 manuscript/manuscript.Rmd
added some text.

@@
 This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)).
-To accomplish these two objectives, we advocate the following four measures: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse.
-It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspires researchers to ask questions that are bold and transformative [@schmidt2020].
+The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+To accomplish these two objectives, we advocate the following four measures as integral components of the Triple P framework: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse.
+It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
 Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.
@@
 Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020].
+The controlled experiments conducted on stone artifacts [@li2022], partcularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
+Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter.
 In the past decades, actualistic experiments becomes more common [@liu2022].
@@
 
-# The ethology and ethnography of stone toolmaking
+# The ethology and ethnography of stone toolmaking experiments
 
+As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)).
 new toolkit such as BORIS were introduced [@friard2016]
 
-```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram of the Perception-Process-Product conceptual framework."}
+```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram demonstrating how to operationalize the Perception-Process-Product conceptual framework."}
 knitr::include_graphics(here::here("figure/Fig1.jpg"))
@@
 
-Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping [@whittaker2004] are not novel, however, it has never been formally recognized as a legitimate research method in experimental archaeology.
+Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel.
+However, it has never been formally recognized as a legitimate research method in experimental archaeology.
 
@@
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
-When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits, preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
+When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
 The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
-However, what is problematic is that the results of replication experiments conducted by these experienced practioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
+However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
 
-Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979].
+It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics. One such example is the the edge angle [@crabtree1977]
+
+Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
 

### 2023-03-18 manuscript/manuscript.Rmd
updated the outline and readme file.

@@
 
-# Introduction
+\vspace{12pt}
 
@@
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
-Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.
+Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.
 
-Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020].
+# What good is actualistic experiment?
+
+Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008].
 The controlled experiments conducted on stone artifacts [@li2022], partcularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 
-# The ethology and ethnography of stone toolmaking experiments
+# The ethology and ethnography of stone toolmaking
 
@@
 
-Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
 
@@
 
+# Acknowledgements
+
+This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" (C. L.).
+
 # References

### 2023-03-20 manuscript/manuscript.Rmd
addded some text from abandoned draft of the Boxgrove paper.

@@
 
+Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.
+Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline (Faniel et al., 2018; Huggett, 2018; Moody et al., 2021).
+Among many reasons preventing archaeologists from reusing published and digitized data (Sobotkova, 2018), the lack of a standardized practice of and motivation for data sharing is a prominent one (Marwick & Birch, 2018).
+As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository.
+Another major and legitimate concern of archaeological data reuse is their quality.
+In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
+Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct
+381 access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
+
 # Acknowledgements

### 2023-03-21 manuscript/manuscript.Rmd
minor changes.

@@
 The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
-To accomplish these two objectives, we advocate the following four measures as integral components of the Triple P framework: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse.
+To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs in a global scale; 5) building an open-access repository for data reuse.
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].

### 2023-03-28 manuscript/manuscript.Rmd
added some references.
minor changes.

@@
 Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008].
-The controlled experiments conducted on stone artifacts [@li2022], partcularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
+The controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
 Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter.
@@
 As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)).
+Ethological approaches are widly used in the description and analysis of non-human animal behavior [@fragaszy2018].
 new toolkit such as BORIS were introduced [@friard2016]
@@
 
-The last step is uploading the data to a open-access repository [@marwick2017]. The building of manufacture can cost [@simon2015; @gilmore2015].
+The last step is uploading the data to a open-access repository [@marwick2017].
+The building of manufacture can cost [@simon2015; @gilmore2015].
+Following the data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020]
 
@@
 In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
-Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct
-381 access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
+Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct 381 access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
 

### 2023-03-29 manuscript/manuscript.Rmd
added some references.

@@
 Experimental archaeology is based on the concept of analogy (i.e., the past is at least partially similar to the present in some aspects).
-It is acknowledged that the validity of this type of analogical inference has long been a subject of debate in archaeology (Wylie, 1985), and a comprehensive review of it is beyond the scope of this essay.
+It is acknowledged that the validity of this type of analogical inference has long been a subject of debate in archaeology [@wylie1985; @chapman2016], and a comprehensive review of it is beyond the scope of this essay.
 For instance, presumably in the paleolithic period the development of stone knapping skills mostly happened during childhood and these children grew up in an environment surrounded by habitual stone tool users and makers.
 In this case, what can knapping teaching experiments involving modern adults who have zero exposure to stone tools inform us about the past learning behavior?
-It is important to clarify that no experimental project ever intends or claims to provide the perfect reconstruction of the past but rather aims at identifying variables relevant to the question of interest (Stout & Khreisheh, 2015), which can be often ignored through pure deductive reasoning.
+It is important to clarify that no experimental project ever intends or claims to provide the perfect reconstruction of the past but rather aims at identifying variables relevant to the question of interest [@stoutSkillLearningHuman2015], which can be often ignored through pure deductive reasoning.
 In the end, all experiments are wrong, but some are useful, and we need more of them.
@@
 As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)).
-Ethological approaches are widly used in the description and analysis of non-human animal behavior [@fragaszy2018].
+Ethological approaches are widely used in the description and analysis of non-human animal behavior [@fragaszy2018].
 new toolkit such as BORIS were introduced [@friard2016]

### 2023-04-20 manuscript/manuscript.Rmd
updated some references.
added some references.

@@
 
-Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008].
+Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
 The controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example.
-Other coding scheme also exist such as [@mahaney2014].
+Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023]
 
@@
 However, it has never been formally recognized as a legitimate research method in experimental archaeology.
+Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
 

### 2023-05-05 manuscript/manuscript.Rmd
more references added.
added some new references.

@@
 In the past decades, actualistic experiments becomes more common [@liu2022].
-Is RCT the golden standard of knowledge [@cartwright2007]
+Is RCT the golden standard of knowledge [@cartwright2007] Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were invovled.
+Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].
 
@@
 Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example.
-Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023]
+Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].
 

### 2023-05-11 manuscript/manuscript.Rmd
added some more references.
minor edits of the first section.

@@
 
-Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
-The controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
-Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter.
-In the past decades, actualistic experiments becomes more common [@liu2022].
-Is RCT the golden standard of knowledge [@cartwright2007] Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were invovled.
+The now extensive corpus of experimental archaeology have principally focusesd on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
+There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
+Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.
+
+1) True controlled experiments are rare.
+Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved.
 Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].
+It has been a debatable issue in philiosophy of science whether RCT represents the golden standard of knowledge [@cartwright2007].
+2) Trade-off between causality and generalizability in experimental design.
+For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.
+
+In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]
 
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
+This also echoes the post-positivist turn in psychology, a field that is dominated by experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
 

### 2023-05-23 manuscript/manuscript.Rmd
minor revision.
added new references.

@@
 
-Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
+Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
 This approach is a power and elegant yet limited by the curse of expertise [@hinds1999].

### 2023-05-25 manuscript/manuscript.Rmd
updated the abstract.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
-abstract: "This paper presents the Perception-Process-Product analytical framework that aims to expand the scope of experimental archaeology .
+abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
+The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product.
+Here we propose the following five basic measures to put the Triple P framework into practice: 1) the acknowledgement of the contribution and limitations of actualistic experiments properly; 2) the normalization the ethological and ethnographic data collection in experimental projects; 3) the involvement of avocational as well as novice participants; 4) the collaboration across labs on a global scale; and 5) the development of an open-access repository for data reuse.
 \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
@@
 The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
-To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs in a global scale; 5) building an open-access repository for data reuse.
+To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse.
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
@@
 
-In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]
+In the past decades, actualistic experiments becomes more common [@liu2022].
+Is RCT the golden standard of knowledge [@cartwright2007].
+Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
 

### 2023-05-25 manuscript/manuscript.Rmd
minor updates.

@@
 
-The now extensive corpus of experimental archaeology have principally focusesd on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
+The now extensive corpus of experimental archaeology have principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
 There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].
-It has been a debatable issue in philiosophy of science whether RCT represents the golden standard of knowledge [@cartwright2007].
+It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].
 2)  Trade-off between causality and generalizability in experimental design.
@@
 In the past decades, actualistic experiments becomes more common [@liu2022].
-Is RCT the golden standard of knowledge [@cartwright2007].
 Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
 
-the first two p captures different level of variation: EQUIFINALITY [@chami2015].
 The process level can be dismantled into two parts: 1)the cognitive [@stout2015] and physical demands [@key2018] of certain tool-making behavior; 2) the ethological analysis.
@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)).
-Ethological approaches are widely used in the description and analysis of non-human animal behavior [@fragaszy2018].
-new toolkit such as BORIS were introduced [@friard2016]
+As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality, one of daunting challenges faced in archaeological inference[@chami2015].
+Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths.
+Ethological approaches are widely used in the description and analysis of non-human animal behavior [@fragaszy2018]. new toolkit such as BORIS were introduced [@friard2016]
 

### 2023-05-30 manuscript/manuscript.Rmd
references updates.

@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality, one of daunting challenges faced in archaeological inference[@chami2015].
-Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths.
+As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality, one of daunting challenges faced in archaeological inference [@hiscock2004; @lake2014; @premoEquifinalityExplanationThoughts2010].
+Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
+
 Ethological approaches are widely used in the description and analysis of non-human animal behavior [@fragaszy2018]. new toolkit such as BORIS were introduced [@friard2016]

### 2023-05-30 manuscript/manuscript.Rmd
minor updates.

@@
 This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
 To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse.
@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality, one of daunting challenges faced in archaeological inference [@hiscock2004; @lake2014; @premoEquifinalityExplanationThoughts2010].
+As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
+While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionalities simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better understand the range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.
 

### 2023-06-02 manuscript/manuscript.Rmd
minor updates.

@@
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionalities simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better understand the range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.
-
-Ethological approaches are widely used in the description and analysis of non-human animal behavior [@fragaszy2018]. new toolkit such as BORIS were introduced [@friard2016]
+While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.
 
@@
 
-Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal.
-This approach is a power and elegant yet limited by the curse of expertise [@hinds1999].
-Like chaine operatoire, it cannot handles variation very well.
-To some extent it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps.
-Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation.
-Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example.
+While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences invovled in the manufacture of material remains are not infrequent.
+One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023].
+Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
+This approach is a power and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handles variability very well.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps, while novices often feature a low planning depth [@vanopheusden2023] and a different sets of perception on the causal structure of how certain behaviors will modify the raw materials.
+
+Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis.
+With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years.
+Here we use action grammar developed by [@stout2021] as an example.
 Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].

### 2023-06-03 manuscript/manuscript_word.Rmd
minor updates.

--- 2023-06-02 manuscript/manuscript.Rmd
@@
 date: "`r Sys.Date()`"
-fontsize: 11pt
-output: bookdown::pdf_document2
-header-includes:
-   - \usepackage{textcomp}
-   - \usepackage{utopia}
-   - \usepackage{lineno}
-   - \linenumbers
-   - \usepackage{float}
-   - \floatplacement{figure}{H}
-   - \usepackage{caption}
-   - \captionsetup[figure]{font=scriptsize}
-   - \captionsetup[table]{font=scriptsize}
+output:
+  word_document
 linestretch: 1.5
@@
 csl: apa.csl
+fontsize: 11pt
+always_allow_html: true
 ---
@@
 This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to a) amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and b) better identify the complex interacting relationships across these three levels of variations.
 To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse.
@@
 
-1)  True controlled experiments are rare.
-Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved.
+True controlled experiments are rare.
+Randomized controlled experiments sensu stricto are extremely rare in experimental archaeology when human participants were involved.
 Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].
 It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].
-2)  Trade-off between causality and generalizability in experimental design.
+
+Trade-off between causality and generalizability in experimental design.
 For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.
@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
+As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (Figure @ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
@@
 
-```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram demonstrating how to operationalize the Perception-Process-Product conceptual framework."}
+{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram demonstrating how to operationalize the Perception-Process-Product conceptual framework."}
 knitr::include_graphics(here::here("figure/Fig1.jpg"))
-```
 

### 2023-06-05 manuscript/manuscript.Rmd
added another figure.

@@
 
-The now extensive corpus of experimental archaeology have principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
+The now extensive corpus of experimental archaeology have witnessed the growing principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
 There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 
-1)  True controlled experiments are rare.
-Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved.
-Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].
-It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].
-2)  Trade-off between causality and generalizability in experimental design.
+Controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
+Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
+Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through expeirments featuring small-sized convenience sample.
+It has also been a debatable issue whether Randomized Controled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
+
+Trade-off between causality and generalizability in experimental design.
 For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.
@@
 Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
-
-The process level can be dismantled into two parts: 1)the cognitive [@stout2015] and physical demands [@key2018] of certain tool-making behavior; 2) the ethological analysis.
+Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
 
@@
 Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis.
-With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years.
+With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
 Here we use action grammar developed by [@stout2021] as an example.
 Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].
+
+```{r ethogram, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="An example of coding Bruce Bradley's handaxe knapping session using the action grammer and BORIS software."}
+knitr::include_graphics(here::here("figure/Fig2.jpg"))
+```
 

### 2023-06-06 manuscript/manuscript.Rmd
minor updates.

@@
 For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.
-
-In the past decades, actualistic experiments becomes more common [@liu2022].
-Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
-Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
-
 One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition.
 The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity.
-Roe and Just (2009: 1266-1267) defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings".
+Roe and Just [@roe2009: 1266-1267] defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings".
 This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs.
-In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades (Eren et al., 2016).
+In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades [@eren2016].
 Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area.
@@
 As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.
+
+In the past decades, actualistic experiments becomes more common [@liu2022].
+Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
+Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
 
@@
 
-While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences invovled in the manufacture of material remains are not infrequent.
+While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
 One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023].
@@
 In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
-Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct 381 access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
+Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
 

### 2023-06-07 manuscript/manuscript.Rmd
minor updates.

@@
 This approach is a power and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handles variability very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps, while novices often feature a low planning depth [@vanopheusden2023] and a different sets of perception on the causal structure of how certain behaviors will modify the raw materials.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different sets of perception on the causal structure of how certain behaviors will modify the raw materials.
 
@@
 
-We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+Contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].

### 2023-06-25 manuscript/manuscript.Rmd
minor updates.

@@
 Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
-Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through expeirments featuring small-sized convenience sample.
-It has also been a debatable issue whether Randomized Controled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
+Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
+It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
 
@@
 In the past decades, actualistic experiments becomes more common [@liu2022].
-Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
+Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018].
 Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
+It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020].
 

### 2023-06-30 manuscript/manuscript.Rmd
minor updates.

@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
+As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
-This also echoes the post-positivist turn in psychology, a field that is dominated by experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
+This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
 

### 2023-07-01 manuscript/manuscript.Rmd
minor updates.

@@
 It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020].
+how decontexualized law-like statements interact in a particular context.
 
@@
 In the end, all experiments are wrong, but some are useful, and we need more of them.
+
+Again, actualistic expeirment as a huestric for hypothesis generation as well as a low-cost , echoing the spirit stated by Lin et al.
+[-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction" [].
 

### 2023-07-04 manuscript/manuscript.Rmd
minor updates.

--- 2023-06-30 manuscript/manuscript.Rmd
@@
 
-The now extensive corpus of experimental archaeology have witnessed the growing principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
+The now extensive corpus of experimental archaeology have witnessed the growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
 There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 
-Trade-off between causality and generalizability in experimental design.
-For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.
-One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition.
-The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity.
-Roe and Just [@roe2009: 1266-1267] defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings".
-This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs.
-In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades [@eren2016].
+Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms another major issue in experimental design [@eren2016; @roe2009: 1266-1267].
+In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades [].
 Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area.
@@
 Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions.
-In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in teaching experiments.
+In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in in a series of recently published experimental studies aiming at understanding the transmission of lithic technologies.
 As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.
@@
 In the end, all experiments are wrong, but some are useful, and we need more of them.
+
+Again, actualistic expeirment as a huestric for hypothesis generation as well as a low-cost , echoing the spirit stated by Lin et al.
+[-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".
 

### 2023-07-04 manuscript/manuscript.Rmd
reorganization of section one.

@@
 
-The now extensive corpus of experimental archaeology have witnessed the growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
+In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
 There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 
-Controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
+Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
+In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped identifying multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapbale of inferring how decontexualized law-like statements interact in a particular context as reflected in real-world conditions.
+In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023].
+In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acqusition.
+After all, these experimental results can only be as robust as its experimental settings.
+
+More importantly, controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
 Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
@@
 
-Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms another major issue in experimental design [@eren2016; @roe2009: 1266-1267].
-In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades [].
-Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area.
-All the variables of interest are easy to measure, quantify, and control in a machine knapping setting.
-Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions.
-In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in in a series of recently published experimental studies aiming at understanding the transmission of lithic technologies.
-As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.
-
-In the past decades, actualistic experiments becomes more common [@liu2022].
-Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018].
-Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
+On the other hand, actualistic experiments pay more attention to how experimental results can be generated to archaeological samples.
+Unlike controlled experiment, variability could be easily observed in actualistic experiment by design, and it cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018].
+Furthermore, statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
 It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020].
-how decontexualized law-like statements interact in a particular context.
-
-Experimental archaeology is based on the concept of analogy (i.e., the past is at least partially similar to the present in some aspects).
-It is acknowledged that the validity of this type of analogical inference has long been a subject of debate in archaeology [@wylie1985; @chapman2016], and a comprehensive review of it is beyond the scope of this essay.
-For instance, presumably in the paleolithic period the development of stone knapping skills mostly happened during childhood and these children grew up in an environment surrounded by habitual stone tool users and makers.
-In this case, what can knapping teaching experiments involving modern adults who have zero exposure to stone tools inform us about the past learning behavior?
-It is important to clarify that no experimental project ever intends or claims to provide the perfect reconstruction of the past but rather aims at identifying variables relevant to the question of interest [@stoutSkillLearningHuman2015], which can be often ignored through pure deductive reasoning.
-In the end, all experiments are wrong, but some are useful, and we need more of them.
-
-Again, actualistic expeirment as a huestric for hypothesis generation as well as a low-cost , echoing the spirit stated by Lin et al.
+Lastly, actualistic experiment can act as a hurestric for hypothesis generation, echoing the spirit stated by Lin et al.
 [-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".
@@
 
-Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020] and museum collection studies [@timbrell2022] of stone artifacts.
+Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
 
@@
 
+# Conclusion
+
+The Triple P framework is theoretically compatible with
+
 # Acknowledgements

### 2023-07-05 manuscript/manuscript.Rmd
finished the first section.

@@
 
-In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
+In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
 There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
@@
 Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
-In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped identifying multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
-All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapbale of inferring how decontexualized law-like statements interact in a particular context as reflected in real-world conditions.
+In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped mapping out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
 In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023].
-In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acqusition.
+In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition.
 After all, these experimental results can only be as robust as its experimental settings.
@@
 
-On the other hand, actualistic experiments pay more attention to how experimental results can be generated to archaeological samples.
-Unlike controlled experiment, variability could be easily observed in actualistic experiment by design, and it cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018].
-Furthermore, statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023].
-It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020].
-Lastly, actualistic experiment can act as a hurestric for hypothesis generation, echoing the spirit stated by Lin et al.
-[-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".
+On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certian degree of compromised control [@outram2008].
+Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
+Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment.
+Unlike controlled experiment, variability could be easily observed in actualistic experiment by design.
+This feature is crutial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018].
+Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023].
+Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
+[-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
 
@@
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.
+While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999].
+The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
 
@@
 
-It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics. One such example is the the edge angle [@crabtree1977]
-
-Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' efforts of reverse engineering during the 19th century, rather than from the intergenerational transmission of knapping knowledge spanning millions of years.
+This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics.
+One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
+Crabtree demonstrated how , challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking.
+Li
 

### 2023-07-06 manuscript/manuscript.Rmd
finished the first draft.

@@
    - \linenumbers
+   - \usepackage[draft]{hyperref}
    - \usepackage{float}
@@
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
-Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.
+Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.
 
@@
 
-In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
-There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
+In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
+There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights into our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
 Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.
 
-Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
-In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped mapping out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
+In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
@@
 In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition.
-After all, these experimental results can only be as robust as its experimental settings.
+After all, these experimental results can only be as robust as their experimental settings.
 
-More importantly, controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
+More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
 Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
-Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
-It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
+Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
+It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
 
-On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certian degree of compromised control [@outram2008].
+On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
 Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
 Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment.
-Unlike controlled experiment, variability could be easily observed in actualistic experiment by design.
-This feature is crutial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018].
+Unlike controlled experiments, variability could be easily observed in actualistic experiments by design.
+This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018].
 Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023].
@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
+As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
@@
 Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
-This approach is a power and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handles variability very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different sets of perception on the causal structure of how certain behaviors will modify the raw materials.
+This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different set of perception on the causal structure of how certain behaviors will modify the raw materials.
 
-Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis.
+Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis.
 With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
-Here we use action grammar developed by [@stout2021] as an example.
-Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].
+Here we use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019].
+The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.
 
@@
 knitr::include_graphics(here::here("figure/Fig2.jpg"))
+```
+
+```{r tab1, echo=FALSE, message=FALSE}
+library(tidyverse)
+df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change"), 
+                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip.
+Often occurs during core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand.
+Often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone", "Exchange of one percussor for another"), check.names=FALSE)
+
+kableExtra::kbl(df1, booktabs = T, caption = "A modiefied version of the original action grammer presented in [@stout2021]", format = "markdown")
 ```
@@
 
+Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities.
+This approach enables researchers to gain a deeper understanding of how people in ancient societies might have perceived and interpreted their environment, objects, and actions.
+Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
+One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+.
+
 # The curse of knowledge
 
-Contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
@@
 
-It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' efforts of reverse engineering during the 19th century, rather than from the intergenerational transmission of knapping knowledge spanning millions of years.
+It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
 This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
-It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics.
+Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
+They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
 One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
-Crabtree demonstrated how , challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking.
-Li
+In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
+It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
+Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
 
@@
 
-Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
+By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts.
+This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
+It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
+This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
 
-In addition to the difficulty in coordination and logistics, the facilitation of large-scale collaborations is often hindered by the current system of research evaluation, where usually only the first author and the senior (last/correspondent) author of a peer-reviewed journal paper will be acknowledged as proper contribution.
+However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
+The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
+This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
 
@@
 
-The last step is uploading the data to a open-access repository [@marwick2017].
-The building of manufacture can cost [@simon2015; @gilmore2015].
-Following the data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020]
+The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
+Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, while videos are the main vector of process-level data, featuring the highest file size compared with the other two.
+As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
 
-Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.
-Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline (Faniel et al., 2018; Huggett, 2018; Moody et al., 2021).
-Among many reasons preventing archaeologists from reusing published and digitized data (Sobotkova, 2018), the lack of a standardized practice of and motivation for data sharing is a prominent one (Marwick & Birch, 2018).
-As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository.
-Another major and legitimate concern of archaeological data reuse is their quality.
-In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
-Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
+The accessibility and availability of experimental data foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data.
+More importantly, a centralized database also promotes data preservation and long-term accessibility.
+By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time.
+This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments.
+It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated.
+However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
+Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
 
@@
 
-The Triple P framework is theoretically compatible with
+Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variability.
+It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variability) to post-processual archaeology (emphasis on perception through ethnography).
+In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.
+The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has a huge potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
 

### 2023-07-06 manuscript/manuscript_word.Rmd
finished the first draft.

--- 2023-06-02 manuscript/manuscript.Rmd
@@
 
-The now extensive corpus of experimental archaeology have principally focused on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999].
-There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
+In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
+There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights into our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
 Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.
 
-1)  True controlled experiments are rare.
-Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved.
-Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].
-It has also been a debatable issue in philosophy of science whether Randomized Controled Trial (RCT) represents the golden standard of knowledge [@cartwright2007].
-2)  Trade-off between causality and generalizability in experimental design.
-For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.
+The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
+In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
+In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023].
+In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition.
+After all, these experimental results can only be as robust as their experimental settings.
 
-In the past decades, actualistic experiments becomes more common [@liu2022].
-Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [@stout2002; @arthur2018].
+More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
+Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
+Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
+It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
 
-The process level can be dismantled into two parts: 1)the cognitive [@stout2015] and physical demands [@key2018] of certain tool-making behavior; 2) the ethological analysis.
-
-One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition.
-The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity.
-Roe and Just (2009: 1266-1267) defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings".
-This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs.
-In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades (Eren et al., 2016).
-Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area.
-All the variables of interest are easy to measure, quantify, and control in a machine knapping setting.
-Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions.
-In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in teaching experiments.
-As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.
-
-Experimental archaeology is based on the concept of analogy (i.e., the past is at least partially similar to the present in some aspects).
-It is acknowledged that the validity of this type of analogical inference has long been a subject of debate in archaeology [@wylie1985; @chapman2016], and a comprehensive review of it is beyond the scope of this essay.
-For instance, presumably in the paleolithic period the development of stone knapping skills mostly happened during childhood and these children grew up in an environment surrounded by habitual stone tool users and makers.
-In this case, what can knapping teaching experiments involving modern adults who have zero exposure to stone tools inform us about the past learning behavior?
-It is important to clarify that no experimental project ever intends or claims to provide the perfect reconstruction of the past but rather aims at identifying variables relevant to the question of interest [@stoutSkillLearningHuman2015], which can be often ignored through pure deductive reasoning.
-In the end, all experiments are wrong, but some are useful, and we need more of them.
+On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
+Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
+Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment.
+Unlike controlled experiments, variability could be easily observed in actualistic experiments by design.
+This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018].
+Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023].
+Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
+[-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
 
@@
 
-As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
+As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.
+While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999].
+The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
 
@@
 
-While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences invovled in the manufacture of material remains are not infrequent.
+While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
 One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023].
 Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
-This approach is a power and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handles variability very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps, while novices often feature a low planning depth [@vanopheusden2023] and a different sets of perception on the causal structure of how certain behaviors will modify the raw materials.
+This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different set of perception on the causal structure of how certain behaviors will modify the raw materials.
 
-Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conduct systematical ethogram analysis.
-With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years.
-Here we use action grammar developed by [@stout2021] as an example.
-Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023], or [@cueva-temprana2019].
+Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis.
+With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
+Here we use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019].
+The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.
+
+```{r ethogram, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="An example of coding Bruce Bradley's handaxe knapping session using the action grammer and BORIS software."}
+knitr::include_graphics(here::here("figure/Fig2.jpg"))
+```
+
+```{r tab1, echo=FALSE, message=FALSE}
+library(tidyverse)
+df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change"), 
+                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip.
+Often occurs during core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand.
+Often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone", "Exchange of one percussor for another"), check.names=FALSE)
+
+kableExtra::kbl(df1, booktabs = T, caption = "A modiefied version of the original action grammer presented in [@stout2021]", format = "markdown")
+```
 
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
-This also echoes the post-positivist turn in psychology, a field that is dominated by experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
+This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
+
+Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities.
+This approach enables researchers to gain a deeper understanding of how people in ancient societies might have perceived and interpreted their environment, objects, and actions.
+Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
+One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+.
 
@@
 
-We believe that contemporary practices in experimental archaeology, as manifested by the fact the the the majority of scholarly publications are produced as results of experiments conducted by single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
@@
 
-It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics. One such example is the the edge angle [@crabtree1977]
-
-Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
+This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
+They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
+One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
+In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
+It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
+Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
 
@@
 
-Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020] and museum collection studies [@timbrell2022] of stone artifacts.
+Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
+By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts.
+This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
+It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
+This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
 
-In addition to the difficulty in coordination and logistics, the facilitation of large-scale collaborations is often hindered by the current system of research evaluation, where usually only the first author and the senior (last/correspondent) author of a peer-reviewed journal paper will be acknowledged as proper contribution.
+However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
+The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
+This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
 
@@
 
-The last step is uploading the data to a open-access repository [@marwick2017].
-The building of manufacture can cost [@simon2015; @gilmore2015].
-Following the data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020]
+The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
+Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, while videos are the main vector of process-level data, featuring the highest file size compared with the other two.
+As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
 
-Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.
-Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline (Faniel et al., 2018; Huggett, 2018; Moody et al., 2021).
-Among many reasons preventing archaeologists from reusing published and digitized data (Sobotkova, 2018), the lack of a standardized practice of and motivation for data sharing is a prominent one (Marwick & Birch, 2018).
-As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository.
-Another major and legitimate concern of archaeological data reuse is their quality.
-In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
-Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct 381 access to archaeological artifacts (Balandier et al., 2022; Ogundiran, 2021).
+The accessibility and availability of experimental data foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data.
+More importantly, a centralized database also promotes data preservation and long-term accessibility.
+By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time.
+This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments.
+It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated.
+However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
+Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
+
+# Conclusion
+
+Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variability.
+It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variability) to post-processual archaeology (emphasis on perception through ethnography).
+In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.
+The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has a huge potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
 

### 2023-07-12 manuscript/manuscript.Rmd
paper reorganization.

@@
 The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product.
-Here we propose the following five basic measures to put the Triple P framework into practice: 1) the acknowledgement of the contribution and limitations of actualistic experiments properly; 2) the normalization the ethological and ethnographic data collection in experimental projects; 3) the involvement of avocational as well as novice participants; 4) the collaboration across labs on a global scale; and 5) the development of an open-access repository for data reuse.
+Here I propose the following five basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
@@
 Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
-To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse.
+To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
@@
 
-# The ethology and ethnography of stone toolmaking
+# Many places, many voices
+
+Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
+When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
+The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
+However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
+
+It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
+This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
+They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
+One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
+In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
+It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
+Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
+
+Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
+By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts.
+This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
+It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
+This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
+
+However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
+The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
+This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
+
+# The Triple P framework in action
 
@@
 
+## Product-level data
+
+## Process-level data
+
 While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
@@
 With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
-Here we use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019].
+Here I use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019].
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
@@
 
+## Perception-level data
+
 Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel.
@@
 One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
-.
 
-# The curse of knowledge
-
-Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
-The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
-When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
-The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
-However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
-
-It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
-This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
-Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
-They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
-One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
-In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
-It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
-Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
-Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
-
-# Many places, many voices
-
-Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
-The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
-By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts.
-This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
-Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
-It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
-By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
-This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
-
-However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
-The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
-This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
-To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
-
-# Open science beyond reproducibility
+## Multi-level data curation
 

### 2023-07-19 manuscript/manuscript.Rmd
added a few references.

@@
 Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
-Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment.
+Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
 Unlike controlled experiments, variability could be easily observed in actualistic experiments by design.

### 2023-07-20 manuscript/manuscript.Rmd
yaml reformatting.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
-abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
+abstract: |
+  ```{=latex}
+  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product.
 Here I propose the following five basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
-\\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
+
+  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production
+  ```
 date: "`r Sys.Date()`"
@@
    - \linenumbers
-   - \usepackage[draft]{hyperref}
    - \usepackage{float}
@@
 
-\vspace{12pt}
+# Introduction
 

### 2023-07-20 manuscript/manuscript.Rmd
reference and acknowledgement updates.

@@
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
-When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
+When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654].
 The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
-However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
+However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
 
-It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
+Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
 This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
@@
 The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
-By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts.
+By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164].
 This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
@@
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999].
+While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023].
 The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
-This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
+This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
 
@@
 
-This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" (C. L.).
+I would like to thank Dietrich Stout for his helpful comments on earlier drafts of this article.
+I am also grateful for my collaborators who help me turn this idea into reality: Russell Cutts (Oxford College of Emory University), Nada Khreisheh (The Ancient Technology Centre), Justin Pargeter (New York University), Ron Shimelmitz (University of Haifa), and particularly Mark Moore (University of New England).
+This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach"
 

### 2023-07-22 manuscript/manuscript.Rmd
minor fix

@@
 I am also grateful for my collaborators who help me turn this idea into reality: Russell Cutts (Oxford College of Emory University), Nada Khreisheh (The Ancient Technology Centre), Justin Pargeter (New York University), Ron Shimelmitz (University of Haifa), and particularly Mark Moore (University of New England).
-This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach"
+This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach."
 

### 2023-08-11 manuscript/manuscript.Rmd
minor updates.

@@
 Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
-It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
+It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].
 
@@
 The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
-Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, while videos are the main vector of process-level data, featuring the highest file size compared with the other two.
+Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
 As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

### 2023-08-14 manuscript/manuscript.Rmd
revision on introduction.

@@
 
-This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology.
+Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists.
+
+As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
 To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
@@
 This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different set of perception on the causal structure of how certain behaviors will modify the raw materials.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
+However, this is not a solid assumption given that 1) experiments and modeling studies in congnitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practioniers in tradional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].
 

### 2023-08-15 manuscript/manuscript.Rmd
1st round revision done based on Dietz's comments.

@@
 This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology.
-Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists.
+Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].
 
@@
 
-In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
-There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights into our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on.
-Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.
-
-The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
+The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267], and the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
 In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
@@
 More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
+This issue is rarely discussed in material science oriented experiments but should raise a critical concern in experimental archaeology as social science.
 Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
@@
 This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018].
-Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023].
+Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007].
 Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
-[-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
+[-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
 
@@
 
-Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
 The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
@@
 
+Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials.
+It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage.
+No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
+Besides, adopting good habits in spreadsheet data organization is strongly recommended [@broman2018].
+
 ## Process-level data
@@
 To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
-However, this is not a solid assumption given that 1) experiments and modeling studies in congnitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practioniers in tradional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].
+However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].
 
@@
 Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities.
-This approach enables researchers to gain a deeper understanding of how people in ancient societies might have perceived and interpreted their environment, objects, and actions.
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
@@
 As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
-
-The accessibility and availability of experimental data foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data.
-More importantly, a centralized database also promotes data preservation and long-term accessibility.
-By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time.
-This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments.
-It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated.
-However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
-Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
 

### 2023-08-15 manuscript/manuscript_word.Rmd
1st round revision done based on Dietz's comments.

@@
 
-\vspace{12pt}
+# Introduction
 
-This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
-To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse.
+This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology.
+Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].
+
+As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
@@
 More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
+This issue is rarely discussed in material science oriented experiments but should raise a critical concern in experimental archaeology as social science.
 Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
 Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
-It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].
+It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].
 
@@
 Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
-Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment.
+Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
 Unlike controlled experiments, variability could be easily observed in actualistic experiments by design.
 This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018].
-Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023].
+Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007].
 Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
-[-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
+[-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
 
-# The ethology and ethnography of stone toolmaking
+# Many places, many voices
+
+Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
+When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654].
+The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
+However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
+
+Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
+This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
+They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
+One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
+In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
+It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
+Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
+
+Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
+By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164].
+This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
+It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
+This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
+
+However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
+The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
+This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
+
+# The Triple P framework in action
 
@@
 
+## Product-level data
+
+Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials.
+It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage.
+No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
+Besides, adopting good habits in spreadsheet data organization is strongly recommended [@broman2018].
+
+## Process-level data
+
 While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
@@
 This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different set of perception on the causal structure of how certain behaviors will modify the raw materials.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
+However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].
 
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
-This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
+This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
 
 Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities.
-This approach enables researchers to gain a deeper understanding of how people in ancient societies might have perceived and interpreted their environment, objects, and actions.
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
 One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
-.
 
-# The curse of knowledge
-
-Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
-The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
-When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable.
-The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
-However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
-
-It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
-This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
-Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
-They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
-One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
-In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
-It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
-Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
-Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
-
-# Many places, many voices
-
-Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
-The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
-By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts.
-This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
-Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
-It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
-By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
-This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
-
-However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
-The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
-This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
-To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
-
-# Open science beyond reproducibility
+## Multi-level data curation
 
 The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
-Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, while videos are the main vector of process-level data, featuring the highest file size compared with the other two.
+Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
 As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
-
-The accessibility and availability of experimental data foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data.
-More importantly, a centralized database also promotes data preservation and long-term accessibility.
-By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time.
-This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments.
-It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated.
-However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
-Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
 
@@
 
-This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" (C. L.).
+I would like to thank Dietrich Stout for his helpful comments on earlier drafts of this article.
+I am also grateful for my collaborators who help me turn this idea into reality: Russell Cutts (Oxford College of Emory University), Nada Khreisheh (The Ancient Technology Centre), Justin Pargeter (New York University), Ron Shimelmitz (University of Haifa), and particularly Mark Moore (University of New England).
+This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach."
 

### 2023-09-07 manuscript/manuscript.Rmd
abstract minor change.

@@
 The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product.
-Here I propose the following five basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
+Here I propose the following three basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 

### 2023-09-07 manuscript/manuscript.Rmd
title and terminology update.

@@
 ---
-title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
+title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
 author: 
@@
   This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product.
+The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
 Here I propose the following three basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
@@
 Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].
 
-As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
 To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
@@
 Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
-Unlike controlled experiments, variability could be easily observed in actualistic experiments by design.
+Unlike controlled experiments, variation could be easily observed in actualistic experiments by design.
 This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018].
@@
 
-Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+Emphasizing variation at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
 The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
@@
 Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
-This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well.
+This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variation very well.
 To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
@@
 
-The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
+The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
 Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
@@
 
-Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variability.
-It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variability) to post-processual archaeology (emphasis on perception through ethnography).
+Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variation.
+It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variation) to post-processual archaeology (emphasis on perception through ethnography).
 In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.

### 2023-09-11 manuscript/manuscript.Rmd
1st trial of 2nd round revision.

@@
 author: 
-- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
+- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
 abstract: |
   ```{=latex}
-  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
+  This paper presents the outline of the Perception-Process-Product ("Triple P") conceptual framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
-Here I propose the following three basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
+Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 
@@
 Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
 
-As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
-To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
+Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework.
+Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments and 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices.
+These two principles are developed to advocate a pluralistic approach to the explanation of complex variation instead of treating the optimization-based research agenda as a panacea.
+Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
@@
 
-# What good is actualistic experiment?
+# What good is actualistic experimentation?
 
-The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267], and the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
+The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
+The past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
 In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
-In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023].
-In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition.
+In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be potentially problematic [@liu2023].
+In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of cultural transmission.
 After all, these experimental results can only be as robust as their experimental settings.
-
-More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012].
-This issue is rarely discussed in material science oriented experiments but should raise a critical concern in experimental archaeology as social science.
-Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved.
-Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample.
-It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].
 
@@
 
-Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
-This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61], rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
 Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
@@
 
-As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265].
+As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
@@
 No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
-Besides, adopting good habits in spreadsheet data organization is strongly recommended [@broman2018].
+Adopting good habits in spreadsheet data organization is also strongly recommended [@broman2018].
 
@@
 
-While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
+While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
 One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023].
 Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
-This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variation very well.
+This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well.
 To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
@@
 With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
-Here I use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019].
+Here I use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023] or granularity [@mahaney2014; @cueva-temprana2019; @roux2005].
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
@@
 
-Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel.
-However, it has never been formally recognized as a legitimate research method in experimental archaeology.
-Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
+Ethnographies revolving around experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S.
+hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel.
+However, ethnography has never been formally recognized as a legitimate research method in experimental archaeology.
+Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
 This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
 
-Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities.
+Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
@@
 
-The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
+The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology.
 Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.

### 2023-09-11 manuscript/manuscript_word.Rmd
1st trial of 2nd round revision.

@@
 ---
-title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
+title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
 author: 
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
-abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product.
-Here I propose the following five basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
-\\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
+abstract: "Abstract: This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
+The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
+Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
+\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
 date: "`r Sys.Date()`"
@@
 Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
 
-As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
-To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
+Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework.
+Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments and 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices.
+These two principles are developed to advocate a pluralistic approach to the explanation of complex variation instead of treating the optimization-based research agenda as a panacea.
+Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
 It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
@@
 
-While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
+While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
 One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023].
 Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
-This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well.
+This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well.
 To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
@@
 
-The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future.
+The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology.
 Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.

### 2023-09-12 manuscript/manuscript.Rmd
finished the 2nd round of revision.

@@
   ```{=latex}
-  This paper presents the outline of the Perception-Process-Product ("Triple P") conceptual framework that aims to expand the scope of experimental archaeology.
+  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
-Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
+Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 
-  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production
+  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production
   ```
@@
 
-Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations.
+Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework.
-Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments and 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices.
-These two principles are developed to advocate a pluralistic approach to the explanation of complex variation instead of treating the optimization-based research agenda as a panacea.
-Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
-It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
+Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability and 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
+These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea.
+Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+There is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
 Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.
@@
 
-The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267].
-The past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
+The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
+In fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
+In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
 In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
-All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
-In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be potentially problematic [@liu2023].
-In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of cultural transmission.
+All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
+In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies [@liu2023].
+In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022].
 After all, these experimental results can only be as robust as their experimental settings.
@@
 Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
-Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
 Unlike controlled experiments, variation could be easily observed in actualistic experiments by design.
-This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018].
-Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007].
+This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
+Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007].
 Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
@@
 Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
-The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989].
+This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989], but it has further implications for the sample representativeness in experimental archaeology.
 When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654].
-The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists.
+The existence of this cognitive bias is not inherently bad, and these many years of experience should be appreciated and celebrated by experimental archaeologists.
 However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
 
-Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61], rather than from the inter-generational transmission of knapping knowledge spanning millions of years.
+Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
 Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
 They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
-One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
+One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree [-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
 In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
@@
 
-Emphasizing variation at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
-The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders.
-By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164].
-This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
-Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise.
-It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
-By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
+Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+This emphasis on team efforts and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
 This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
-
-However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
-The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
-This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
-To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
 
@@
 
-While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
-One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023].
-Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal.
+While systematic behavioral coding methods that are widely used in the study of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
+One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @haidle2014; @haidle2023].
+Cognigram is a graphical representation of the reconstructed behavior behind archaeological artifacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
 This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well.

### 2023-09-12 manuscript/manuscript_word.Rmd
finished the 2nd round of revision.

@@
 author: 
-- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
+- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
 abstract: "Abstract: This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
-Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.
-\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
+Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production"
 date: "`r Sys.Date()`"

### 2023-09-13 manuscript/manuscript.Rmd
minor updates in introduction.

@@
 Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
+In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation [e.g., Sylvia's Recipe, @gergely2006] together with path dependence [e.g., QUERTY keyboard design, @kafaee2022] can lead to the widespread and long-lasting reproduction of technological solutions that are neither optimal in terms of functional efficiency nor minimal in terms of manufacture complexity.
 
@@
 Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
-[-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
+[-@lin2018: 680-681], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
 
@@
 
-Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
+Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker2004], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
 This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989], but it has further implications for the sample representativeness in experimental archaeology.

### 2023-09-19 manuscript/manuscript.Rmd
revision for Antiquity submission.

@@
 
-This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology.
+This paper presents the Perception-Process-Product (hereinafter referred to as 'Triple P') conceptual framework to expand the scope of experimental archaeology.
 Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
@@
 
-Built upon the *Homo economicus* critique and early works in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+Built upon the *Homo economicus* critique and early works in behavioural archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioural channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework.
@@
 There is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
-Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.
+Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artefacts to clarify its meaning and demonstrate the necessity and potential of this framework.
 
@@
 
-The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
+The trade-off between causality (aka 'internal validity') and generalizability (aka 'external validity') forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
 In fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
-In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018].
-In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioural level in the explanation of material culture variation [@eren2016; @lin2018].
+In the context of stone artefact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artefact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
@@
 Unlike controlled experiments, variation could be easily observed in actualistic experiments by design.
-This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
-Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007].
+This feature is crucial and cannot be simply replaced by ethnographic records, because many palaeolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
+Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
 Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
@@
 
-Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker2004], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise".
-This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989], but it has further implications for the sample representativeness in experimental archaeology.
+Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker2004], tend to be restrained by the cognitive bias known as the 'curse of knowledge' or 'curse of expertise'.
+This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999], but it has further implications for the sample representativeness in experimental archaeology.
 When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654].
@@
 
-Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @coles1979; @johnson1978; @whittaker1994: 54-61].
 Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
@@
 It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
-Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Likewise, novices' lack of expertise also helps to mitigate the 'curse of knowledge' bias that may hinder expert knappers.
 Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
 
-Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.
+Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artefacts.
 Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
@@
 Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
-This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].
+This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023].
 
@@
 
-As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
+As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
 Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023].
+While we cannot fully solve these two problems and accurately reconstruct the past behavioural processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioural mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023].
 The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
@@
 Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials.
-It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage.
+It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artefact or a whole assemblage.
 No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
@@
 
-While systematic behavioral coding methods that are widely used in the study of non-human animal behavior [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent.
-One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @haidle2014; @haidle2023].
-Cognigram is a graphical representation of the reconstructed behavior behind archaeological artifacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
+While systematic behavioural coding methods that are widely used in the study of non-human animal behaviour [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioural sequences involved in the manufacture of material remains are not infrequent.
+One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014].
+Cognigram is a graphical representation of the reconstructed behaviour behind archaeological artefacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
 This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well.
 To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
-However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].
+However, this may be biased given that ethnographic studies demonstrated that expert practitioners in traditional societies can have a different set of causal perception on how certain behaviours will modify the raw materials [@harris2021].
 
-Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis.
+Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematic ethogram analysis.
 With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
@@
 
-```{r ethogram, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="An example of coding Bruce Bradley's handaxe knapping session using the action grammer and BORIS software."}
+```{r ethogram, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="An example of coding a handaxe knapping session using the BORIS software."}
 knitr::include_graphics(here::here("figure/Fig2.jpg"))
@@
 
-kableExtra::kbl(df1, booktabs = T, caption = "A modiefied version of the original action grammer presented in [@stout2021]", format = "markdown")
+kableExtra::kbl(df1, booktabs = T, caption = "A modified version of the original action grammar presented in [@stout2021]", format = "markdown")
 ```
@@
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
-One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+One potential application of ethnographic methods in experimental archaeology of stone artefacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
 
@@
 
-The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology.
+The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure, which unfortunately still does not exist yet in experimental archaeology.
 Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
-As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
+As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
 
@@
 Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variation.
-It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variation) to post-processual archaeology (emphasis on perception through ethnography).
+It is also compatible with many theoretical orientations, ranging from behavioural archaeology (emphasis on video recording of behavioural processes) through evolutionary archaeology (emphasis on the amplification of variation) to post-processual archaeology (emphasis on perception through ethnography).
 In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.
@@
 
-I would like to thank Dietrich Stout for his helpful comments on earlier drafts of this article.
-I am also grateful for my collaborators who help me turn this idea into reality: Russell Cutts (Oxford College of Emory University), Nada Khreisheh (The Ancient Technology Centre), Justin Pargeter (New York University), Ron Shimelmitz (University of Haifa), and particularly Mark Moore (University of New England).
-This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach."
+I thank Mark Moore for an inspiring discussion and Dietrich Stout for his helpful comments on earlier drafts of this article.
+This study was supported by a research grant from the Leakey Foundation titled 'Inferring skill reproduction from stone artifacts: A middle‐range approach.'
 

### 2023-09-26 manuscript/manuscript.Rmd
3rd round revision.

@@
   This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
-Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product in order to provide a more contextualized and comprehensive explanation of the past archaeological record.
+I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the inherent trade-off between control and generalizability in actualistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 
@@
 
-This paper presents the Perception-Process-Product (hereinafter referred to as 'Triple P') conceptual framework to expand the scope of experimental archaeology.
+This paper presents the Perception-Process-Product (hereafter 'Triple P') conceptual framework to expand the scope of experimental archaeology.
 Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
-In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation [e.g., Sylvia's Recipe, @gergely2006] together with path dependence [e.g., QUERTY keyboard design, @kafaee2022] can lead to the widespread and long-lasting reproduction of technological solutions that are neither optimal in terms of functional efficiency nor minimal in terms of manufacture complexity.
+Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
+In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation [e.g., Sylvia's Recipe, @gergely2006] together with path dependence [e.g., QUERTY keyboard design, @kafaee2022] can lead to the widespread and long-lasting reproduction of technological solutions that are neither optimal in functional efficiency nor minimal in manufacture complexity.
 
@@
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework.
-Ontologically speaking, it requires 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability and 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
+Ontologically speaking, it requires 1) acknowledging the inherent trade-off between control and generalizability in actualistic research design and 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
 These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea.
@@
 
-The trade-off between causality (aka 'internal validity') and generalizability (aka 'external validity') forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
-In fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
+The trade-off between causal inference (aka 'internal validity') and generalization (aka 'external validity') forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
+Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
 In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioural level in the explanation of material culture variation [@eren2016; @lin2018].
-In the context of stone artefact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artefact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+In the context of stone artefact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artefact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
-In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies [@liu2023].
-In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022].
-After all, these experimental results can only be as robust as their experimental settings.
+Similarly standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023].
+In real-world knapping, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022].
 
@@
 This feature is crucial and cannot be simply replaced by ethnographic records, because many palaeolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
-Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
-Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al.
+While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
+Actualistic experiment can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
 [-@lin2018: 680-681], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
@@
 In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis.
-It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography.
+It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] ethnography.
 Likewise, novices' lack of expertise also helps to mitigate the 'curse of knowledge' bias that may hinder expert knappers.
@@
 
-As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
-Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioural processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioural mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023].
+The implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
+Equifinality refers situation in which a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
+While we cannot fully solve these two problems and accurately reconstruct the past behavioural processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby draw a more informed inference of certain behavioural mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023].
 The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
@@
 Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials.
-It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artefact or a whole assemblage.
+It can exist in the form of spreadsheets containing detailed technological attributes, photos and illustrations, or high-resolution 3D scans of individual artefact or a whole assemblage.
 No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
@@
 
-While systematic behavioural coding methods that are widely used in the study of non-human animal behaviour [@fragaszy2018] still largely neglected among archaeologists, the attempts of reconstructing behavioural sequences involved in the manufacture of material remains are not infrequent.
-One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014].
-Cognigram is a graphical representation of the reconstructed behaviour behind archaeological artefacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
-This approach is a powerful and elegant yet limited by its normative and analytical orientation, meaning it cannot handle variation very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner.
-However, this may be biased given that ethnographic studies demonstrated that expert practitioners in traditional societies can have a different set of causal perception on how certain behaviours will modify the raw materials [@harris2021].
+While systematic behavioural coding methods that are widely used in the study of non-human animal behaviour [@fragaszy2018] are still largely neglected among archaeologists, attempts of reconstructing behavioural sequences involved in the manufacture of material remains are not infrequent.
+One such example is the cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014].
+A cognigram is a graphical representation of the reconstructed behaviour behind archaeological artefacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
+This approach provides an elegant descriptive methodology yet is limited by its normative and analytical orientation, meaning it cannot handle variation very well.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes uniform causal thinking between each step in an idealistic manner.
+However, this may be biased given that ethnographic studies demonstrated that expert non-academic practitioners can have a different set of causal understanding [@harris2021].
 
@@
 However, ethnography has never been formally recognized as a legitimate research method in experimental archaeology.
-Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
+Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
 This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
@@
 The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure, which unfortunately still does not exist yet in experimental archaeology.
-Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
+Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
 As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
@@
 
-# Acknowledgements
+# Acknowledgments
 
-I thank Mark Moore for an inspiring discussion and Dietrich Stout for his helpful comments on earlier drafts of this article.
+I thank Dietrich Stout for his helpful comments on earlier drafts of this article and Mark Moore for an inspiring discussion.
 This study was supported by a research grant from the Leakey Foundation titled 'Inferring skill reproduction from stone artifacts: A middle‐range approach.'

### 2023-09-26 manuscript/manuscript_word.Rmd
3rd round revision.

@@
 abstract: "Abstract: This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes the amplification of multi-level variation and the identification of interacting relationships of variations across the levels of perception, process, and product.
-Here I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments in the trade-off of causality and generalizability; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product in order to provide a more contextualized and comprehensive explanation of the past archaeological record.
+I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the inherent trade-off between control and generalizability in actualistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production"

### 2023-09-27 manuscript/manuscript.Rmd
submitted to antiquity.

@@
   ```{=latex}
-  This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product in order to provide a more contextualized and comprehensive explanation of the past archaeological record.
-I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the inherent trade-off between control and generalizability in actualistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+  This paper presents the Perception-Process-Product (‘Triple P’) framework that aims to expand the scope of experimental archaeology.
+The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
+It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in naturalistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 
@@
 This paper presents the Perception-Process-Product (hereafter 'Triple P') conceptual framework to expand the scope of experimental archaeology.
-Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
+The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+Nevertheless, Occam's razor can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
 In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation [e.g., Sylvia's Recipe, @gergely2006] together with path dependence [e.g., QUERTY keyboard design, @kafaee2022] can lead to the widespread and long-lasting reproduction of technological solutions that are neither optimal in functional efficiency nor minimal in manufacture complexity.
@@
 Built upon the *Homo economicus* critique and early works in behavioural archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioural channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
-To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework.
-Ontologically speaking, it requires 1) acknowledging the inherent trade-off between control and generalizability in actualistic research design and 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
+To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in naturalistic research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
 These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea.
-Epistemologically speaking, the Triple P framework 3) adopts a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
-There is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
-Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artefacts to clarify its meaning and demonstrate the necessity and potential of this framework.
+The second principle particularly allows researchers to develop research questions that are also meaningful to descendant communities through respectful conversation and collaboration [@montgomery2023].
+The Triple P framework also **3)** adopts a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+It is acknowledged that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
+Here I will leverage the extensive corpus in experimental designs and inferences revolving around stone artefacts to clarify its meaning and demonstrate the necessity and potential of this framework.
 
-# What good is actualistic experimentation?
+# What good is naturalistic experimentation?
 
@@
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
-Similarly standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023].
+Similarly, standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023].
 In real-world knapping, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022].
 
-On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
-Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
-Unlike controlled experiments, variation could be easily observed in actualistic experiments by design.
+On the other hand, naturalistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
+Back to the cases of cultural transmission experiments, a naturalistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
+Unlike controlled experiments, variation could be easily observed in naturalistic experiments by design.
 This feature is crucial and cannot be simply replaced by ethnographic records, because many palaeolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
-While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by actualistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
-Actualistic experiment can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
-[-@lin2018: 680-681], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."
+While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by naturalistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
+Naturalistic experiment can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
+[-@lin2018: 680-681], who proposed that the interaction between naturalistic and controlled experiment "operates in a cyclical form of induction and deduction."
 
@@
 The existence of this cognitive bias is not inherently bad, and these many years of experience should be appreciated and celebrated by experimental archaeologists.
-However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.
+However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a vast range of technological diversity.
 
@@
 Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
-They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics.
+They tend to generate ideas that may appear to be counter-intuitive at first glance for academics.
 One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree [-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology.
@@
 Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
-This emphasis on team efforts and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
 Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
@@
 The implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
-Equifinality refers situation in which a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioural processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby draw a more informed inference of certain behavioural mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023].
-The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
+Equifinality refers to situations in which a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
+While we cannot fully solve these two problems and accurately reconstruct the past behavioural processes simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation and draw a more informed inference [@reynolds1999].
+The importance of specifying and documenting the context information of both the experiment and the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
 
@@
 Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials.
-It can exist in the form of spreadsheets containing detailed technological attributes, photos and illustrations, or high-resolution 3D scans of individual artefact or a whole assemblage.
-No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
-Adopting good habits in spreadsheet data organization is also strongly recommended [@broman2018].
+It can exist in the form of spreadsheets containing detailed technological attributes, photos and illustrations, or high-resolution 3D scans of individual artefacts or a whole assemblage.
+No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version, etc.) should be always available in the relevant meta-data.
+I also strongly recommend adopting good habits in spreadsheet data organization [@broman2018].
 
@@
 
-While systematic behavioural coding methods that are widely used in the study of non-human animal behaviour [@fragaszy2018] are still largely neglected among archaeologists, attempts of reconstructing behavioural sequences involved in the manufacture of material remains are not infrequent.
-One such example is the cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014].
+While systematic behavioural coding methods widely used in the study of non-human animal behaviour [@fragaszy2018] are still largely neglected among archaeologists, attempts to reconstruct behavioural sequences involved in the manufacture of material remains are not infrequent.
+One such example is the cognigram, which was first systematically developed and applied in archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014].
 A cognigram is a graphical representation of the reconstructed behaviour behind archaeological artefacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
 This approach provides an elegant descriptive methodology yet is limited by its normative and analytical orientation, meaning it cannot handle variation very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes uniform causal thinking between each step in an idealistic manner.
-However, this may be biased given that ethnographic studies demonstrated that expert non-academic practitioners can have a different set of causal understanding [@harris2021].
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and reflects the analyst's own causal perception.
+However, this may be biased because 1) certain causal insights in stone fracture mechanics remain opaque to academic knappers until they are revealed through controlled experiments by Dibble and his colleagues [@li2022] 2) ethnographic studies demonstrated that expert non-academic practitioners can have a different set of causal understanding [@harris2021].
 
-Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematic ethogram analysis.
+Consequently, we need to accumulate more real-world data by recording a large number of toolmaking videos and conducting systematic ethogram analysis.
 With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)).
@@
 df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change"), 
-                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip.
-Often occurs during core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand.
-Often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone", "Exchange of one percussor for another"), check.names=FALSE)
+                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip.", "Rotation and/or reorientation of the core involving repositioning of the hand.", "Abrasion of core edges using a hammerstone", "Exchange of one percussor for another"), check.names=FALSE)
 
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
-This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
+This also echoes the post-positivist turn in psychology in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
 
@@
 In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.
-The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has a huge potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
+The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has enormous potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
 
-# Acknowledgments
+# Acknowledgements
 

### 2023-09-27 manuscript/manuscript_word.Rmd
submitted to antiquity.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
-abstract: "Abstract: This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product in order to provide a more contextualized and comprehensive explanation of the past archaeological record.
-I propose the following three ontologically and epistemologically based principles to put the Triple P framework into practice: 1) acknowledging the inherent trade-off between control and generalizability in actualistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+abstract: "Abstract: This paper presents the Perception-Process-Product (‘Triple P’) framework that aims to expand the scope of experimental archaeology.
+The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
+It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in naturalistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production"

### 2023-11-23 manuscript/manuscript.Rmd
minor updates.

@@
 
-  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production
+  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production
   ```
@@
 library(tidyverse)
-df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change"), 
-                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip.", "Rotation and/or reorientation of the core involving repositioning of the hand.", "Abrasion of core edges using a hammerstone", "Exchange of one percussor for another"), check.names=FALSE)
+df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change", "Winding Up"), 
+                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip, which often occurs during the core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand, which is often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone.
+The abrasion movement should come from at least two different directions.", "Exchange of one percussor for another", "Preparational percussor movements towards the core that do not lead to the detachment of flakes, which can either be in direct contact with cores or not."), check.names=FALSE)
 
@@
 
-The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure, which unfortunately still does not exist yet in experimental archaeology.
-Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two.
-As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.
+The comparative study and large-scale synthesis of variation data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2018], which unfortunately still does not exist yet in experimental archaeology.
+The accessibility and availability of experimental data can foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data.
+Moreover, a centralized database also promotes data preservation and long-term accessibility.
+By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time.
+This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments.
+It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated.
+However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
+
+Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two.
+As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project.
+Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researcher; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
 

### 2023-12-12 manuscript/manuscript.Rmd
introduction updates.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
-abstract: |
-  ```{=latex}
-  This paper presents the Perception-Process-Product (‘Triple P’) framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
-It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in naturalistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
-
-  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production
-  ```
+abstract: 
 date: "`r Sys.Date()`"
@@
 The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
+When applied to the experimental study of ancient craftsmanship, it implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context.
+Nevertheless, Occam's razor, or the law of parsimony, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use.
+The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e.
+the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001].
+Parallelly, Heyes [@heyes2012] also questioned the abuse of parsimony in animal behavioural research and further proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.
+
 In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation [e.g., Sylvia's Recipe, @gergely2006] together with path dependence [e.g., QUERTY keyboard design, @kafaee2022] can lead to the widespread and long-lasting reproduction of technological solutions that are neither optimal in functional efficiency nor minimal in manufacture complexity.

### 2023-12-16 manuscript/manuscript.Rmd
revision after antiquity rejection.

--- 2023-11-23 manuscript/manuscript.Rmd
@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
-abstract: |
+abstract: 
+|
   ```{=latex}
@@
 The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-Nevertheless, Occam's razor can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* for most anthropologists [@henrichSearchHomoEconomicus2001].
-In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation [e.g., Sylvia's Recipe, @gergely2006] together with path dependence [e.g., QUERTY keyboard design, @kafaee2022] can lead to the widespread and long-lasting reproduction of technological solutions that are neither optimal in functional efficiency nor minimal in manufacture complexity.
+When applied to the experimental study of ancient craftsmanship, it implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context.
+Nevertheless, Occam's razor, or the law of parsimony, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use.
+The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001].
+Parallelly, Heyes [@heyes2012] also questioned the abuse of parsimony in animal behavioural research and further proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.
 
-Built upon the *Homo economicus* critique and early works in behavioural archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioural channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation together with path dependence can lead to the widespread and long-lasting reproduction of technological solutions that are neither minimal in manufacture complexity nor optimal in functional efficiency.
+Over-imitation means the copying of actions that are causally irrelevant in a goal-directed action sequence [@lyons2007].
+It is a psychological propensity that was suggested to be uniquely prevalent among humans in inter-species [@horner2005; @clay2018] comparisons and cross-cultural contexts [@nielsen2010; @stengelin2020].
+Gergely and Csisbra [-@gergely2006] introduced 'Sylvia's Recipe' that vividly illustrates this cognitive process in the transmission of technical skills.
+Sylvia is an education researcher who developed a unique way of cooking ham roast by observing her mother during childhood, where she cut both ends of a ham.
+Later in life her mother happened to watch her cooking, where she noticed and questioned the purpose of this step of preparation.
+Sylvia could not answer it and was then told that it was processed that way because her mother did not have a pan that was large enough to cook a ham.
+The commonality of this opaque causal perception has also been demonstrated in a recent study of Hadza bowmakers.
+Harris et al.
+[-@harris2021] found that even experienced bowmakers only possess limited causal knowledge regarding the design and construction of bows according to modern engineering principles, meaning they cannot spell out the mechanical (dis)advantages of many morphological features.
+On the other hand, path dependence also constrained the pursuit of functional optimization, where people are aware of the existence of a more efficient solution but still stick to the older one due to the cost of learning or other reasons.
+One such example is the longevity of QWERTY keyboard design [@kafaee2022].
+This deliberately unergonomic solution was invented in the era of typewriters in order to disperse commonly used letters, preventing the most frequently struck 'hammers' from clashing.
+Yet it is still the most common keyboard design today when such constraint does not exist anymore on modern computer hardware.
+In short, we should acknowledge the existence and variation of many "good-enough" technological solutions featuring various degrees of "redundancy" in real-world contexts, which often represent locally adaptive peaks instead of a global optimum in a multimodal fitness landscape due to multiple constraints and trade-off factors [@mesoudi2008; @bettinger1982].
+
+Built upon the *Homo economicus* critique as well as intellectual principles and practices in behavioural archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioural channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in naturalistic research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
@@
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
+This research orientation essentially prioritize the material science aspect over the social science aspect of experimental archaeology.
 Similarly, standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023].
@@
 Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
-This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
 Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
-This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023].
+This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023; @marshall2002; @douglass2020].
+
+However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation.
+The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors.
+This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
 
@@
 
-The implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference.
+The implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010; @hiscock2004], two daunting challenges in archaeological inference.
 Equifinality refers to situations in which a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
@@
 It can exist in the form of spreadsheets containing detailed technological attributes, photos and illustrations, or high-resolution 3D scans of individual artefacts or a whole assemblage.
-No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version, etc.) should be always available in the relevant meta-data.
+No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
 I also strongly recommend adopting good habits in spreadsheet data organization [@broman2018].
@@
 While systematic behavioural coding methods widely used in the study of non-human animal behaviour [@fragaszy2018] are still largely neglected among archaeologists, attempts to reconstruct behavioural sequences involved in the manufacture of material remains are not infrequent.
-One such example is the cognigram, which was first systematically developed and applied in archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014].
+One such example is the cognigram, which was first systematically developed and applied in archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014; @haidle2023; @haidle2010].
 A cognigram is a graphical representation of the reconstructed behaviour behind archaeological artefacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
@@
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
-Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021; @cristino2010], etc.
 
@@
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
-This also echoes the post-positivist turn in psychology in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
+Within the broader context of burgeoning interest in mixed-method research in contemporary social science [@creswell2017], this also echoes the post-positivist turn in psychology in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].
 
@@
 One potential application of ethnographic methods in experimental archaeology of stone artefacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely 'an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen).' In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography become necessary.
 

### 2023-12-16 manuscript/manuscript_word.Rmd
revision after antiquity rejection.

@@
 It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in naturalistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
-\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographical analysis; Curse of knowledge; Collaborative knowledge production"
+\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"
 date: "`r Sys.Date()`"

### 2023-12-21 manuscript/manuscript.Rmd
references updates.

@@
 This paper presents the Perception-Process-Product (hereafter 'Triple P') conceptual framework to expand the scope of experimental archaeology.
-The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
+The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009; @schmidt2019; @blessing2021], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
 When applied to the experimental study of ancient craftsmanship, it implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context.
@@
 Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
-This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions.
+This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022; @milks2023] across different regions.
 Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged.
@@
 
-Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
+Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.

### 2024-02-05 manuscript/manuscript.Rmd
minor revision.

@@
 The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
-It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in naturalistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 
@@
 Nevertheless, Occam's razor, or the law of parsimony, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use.
-The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001].
+The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001; @apicella2020].
 Parallelly, Heyes [@heyes2012] also questioned the abuse of parsimony in animal behavioural research and further proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.
@@
 [-@harris2021] found that even experienced bowmakers only possess limited causal knowledge regarding the design and construction of bows according to modern engineering principles, meaning they cannot spell out the mechanical (dis)advantages of many morphological features.
-On the other hand, path dependence also constrained the pursuit of functional optimization, where people are aware of the existence of a more efficient solution but still stick to the older one due to the cost of learning or other reasons.
-One such example is the longevity of QWERTY keyboard design [@kafaee2022].
+
+On the other hand, path dependence also constrained the pursuit of functional optimization, where people are implicitly or explicitly aware of the existence of a more efficient solution but still stick to the older one due to the cost of learning, cultural conservatism [@morin2022; @acerbi2009; @ghirlanda2006], or other reasons.
+One such example in the evolution of technology is the longevity of QWERTY keyboard design [@kafaee2022].
 This deliberately unergonomic solution was invented in the era of typewriters in order to disperse commonly used letters, preventing the most frequently struck 'hammers' from clashing.
@@
 Built upon the *Homo economicus* critique as well as intellectual principles and practices in behavioural archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioural channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
-To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in naturalistic research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
+To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
 These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea.
@@
 
-# What good is naturalistic experimentation?
+# What good is less-controlled experimentation?
 
 The trade-off between causal inference (aka 'internal validity') and generalization (aka 'external validity') forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
-Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
+Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
 In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioural level in the explanation of material culture variation [@eren2016; @lin2018].
@@
 
-On the other hand, naturalistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
-Back to the cases of cultural transmission experiments, a naturalistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
-Unlike controlled experiments, variation could be easily observed in naturalistic experiments by design.
+On the other hand, less-controlled experiments, which have been traditionally known as naturalistic or actualistic expeirments [see terminological critiques of @eren2024; @conrad2023], pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
+Back to the cases of cultural transmission experiments, a less-controlled experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
+Unlike strictly controlled experiments testing one variable of interest each time, variation could be easily observed in less-controlled experiments by design.
 This feature is crucial and cannot be simply replaced by ethnographic records, because many palaeolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
-While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by naturalistic experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
+While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by less-controlled experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
 Naturalistic experiment can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
-[-@lin2018: 680-681], who proposed that the interaction between naturalistic and controlled experiment "operates in a cyclical form of induction and deduction."
+[-@lin2018: 680-681], who proposed that the interaction between less-controlled and strictly controlled experiment "operates in a cyclical form of induction and deduction."
 
@@
 I thank Dietrich Stout for his helpful comments on earlier drafts of this article and Mark Moore for an inspiring discussion.
-This study was supported by a research grant from the Leakey Foundation titled 'Inferring skill reproduction from stone artifacts: A middle‐range approach.'
+This study was supported by the Leakey Foundation research grant titled 'Inferring skill reproduction from stone artifacts: A middle‐range approach' as well as the International Society of Human Ethology's Owen Aldis Award project titled 'Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission.'
 

### 2024-02-05 manuscript/manuscript_word.Rmd
minor revision.

@@
 The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
-It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in naturalistic research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
 \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"

### 2024-02-14 manuscript/manuscript.Rmd
revision following Dietz's advice and submitted to advances in archaeological practice.

@@
   ```{=latex}
-  This paper presents the Perception-Process-Product (‘Triple P’) framework that aims to expand the scope of experimental archaeology.
+  This paper presents the Perception-Process-Product ("Triple P") framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
@@
 
-This paper presents the Perception-Process-Product (hereafter 'Triple P') conceptual framework to expand the scope of experimental archaeology.
-The field has long tended to adopt the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009; @schmidt2019; @blessing2021], whether explicitly or implicitly, and center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords.
-When applied to the experimental study of ancient craftsmanship, it implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context.
-Nevertheless, Occam's razor, or the law of parsimony, can be insufficient to infer the preferences of 'irrational' agents possessing incomplete information [@armstrong2018] in tool design and use.
-The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing for optimality) for many anthropologists [@henrichSearchHomoEconomicus2001; @apicella2020].
-Parallelly, Heyes [@heyes2012] also questioned the abuse of parsimony in animal behavioural research and further proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.
+This paper presents the Perception-Process-Product (hereafter "Triple P") conceptual framework to expand the scope of experimental archaeology.
+The field has long tended to adopt the principle of Occam's razor [e.g., @blessing2021; @domínguez-rodrigo2008; @reeves2009; @schmidt2019], whether explicitly or implicitly.
+This assumption acts to center inquiry around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information experimentation affords.
+When applied to the experimental study of ancient craftsmanship, Occam's razor, or the law of parsimony, implies that a technological solution that is simpler to reproduce is more likely to be the one used in the archaeological context.
+This is insufficient to infer the preferences of "irrational" agents possessing incomplete information [@armstrong2018] in tool design and use.
+The two conditions described here provide a better approximation of past humans displaying extensive cultural variation as opposed to the assumption of omniscient *Homo economicus* (i.e., the idea that humans are consistently rational and narrowly self-interested agents pursuing optimality) that has been rejected by many anthropologists [@henrichSearchHomoEconomicus2001; @apicella2020].
+Heyes [@heyes2012] similarly questioned the abuse of parsimony in animal behavioral research and proposed that new observational and experimental studies that allow differential predictions to be tested become necessary when both a simple and a complex mechanism can explain the phenomenon of interest.
 
-In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation together with path dependence can lead to the widespread and long-lasting reproduction of technological solutions that are neither minimal in manufacture complexity nor optimal in functional efficiency.
+In fact, there are several reasons why past technologies may violate “parsimonious” assumptions of minimal manufacture complexity and optimal functional efficiency.
+In the evolution of technology, it is rather common that opaque causal perception and its resulting tendency of over-imitation can lead to the widespread and long-lasting reproduction of technological solutions that are neither minimal in manufacture complexity nor optimal in functional efficiency.
 Over-imitation means the copying of actions that are causally irrelevant in a goal-directed action sequence [@lyons2007].
 It is a psychological propensity that was suggested to be uniquely prevalent among humans in inter-species [@horner2005; @clay2018] comparisons and cross-cultural contexts [@nielsen2010; @stengelin2020].
-Gergely and Csisbra [-@gergely2006] introduced 'Sylvia's Recipe' that vividly illustrates this cognitive process in the transmission of technical skills.
+Gergely and Csisbra [-@gergely2006] introduced "Sylvia's Recipe" that vividly illustrates this cognitive process in the transmission of technical skills.
 Sylvia is an education researcher who developed a unique way of cooking ham roast by observing her mother during childhood, where she cut both ends of a ham.
-Later in life her mother happened to watch her cooking, where she noticed and questioned the purpose of this step of preparation.
-Sylvia could not answer it and was then told that it was processed that way because her mother did not have a pan that was large enough to cook a ham.
+Later in life, her mother happened to watch her cooking, where she noticed and questioned the purpose of this step of preparation.
+Sylvia could not answer it and was then told that it was processed that way because her mother did not have a pan that was large enough to cook a full-sized ham.
 The commonality of this opaque causal perception has also been demonstrated in a recent study of Hadza bowmakers.
@@
 
-On the other hand, path dependence also constrained the pursuit of functional optimization, where people are implicitly or explicitly aware of the existence of a more efficient solution but still stick to the older one due to the cost of learning, cultural conservatism [@morin2022; @acerbi2009; @ghirlanda2006], or other reasons.
+On the other hand, path dependence also constrained the pursuit of functional optimization or simplification of manufacturing procedures.
+In this case, people are implicitly or explicitly aware of the existence of a more efficient solution but still stick to the older one due to the cost of learning, cultural conservatism [@morin2022; @acerbi2009; @ghirlanda2006], or other reasons.
 One such example in the evolution of technology is the longevity of QWERTY keyboard design [@kafaee2022].
-This deliberately unergonomic solution was invented in the era of typewriters in order to disperse commonly used letters, preventing the most frequently struck 'hammers' from clashing.
+This deliberately unergonomic solution was invented in the era of typewriters in order to disperse commonly used letters, preventing the most frequently struck "hammers" from clashing.
 Yet it is still the most common keyboard design today when such constraint does not exist anymore on modern computer hardware.
@@
 
-Built upon the *Homo economicus* critique as well as intellectual principles and practices in behavioural archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioural channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+Built upon this critique of *Homo economicus* as well as intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
@@
 It is acknowledged that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020].
-Here I will leverage the extensive corpus in experimental designs and inferences revolving around stone artefacts to clarify its meaning and demonstrate the necessity and potential of this framework.
+Here I will leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.
 
@@
 
-The trade-off between causal inference (aka 'internal validity') and generalization (aka 'external validity') forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
+The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
 Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
-In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioural level in the explanation of material culture variation [@eren2016; @lin2018].
-In the context of stone artefact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artefact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
+In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024].
+In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
 All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions.
-This research orientation essentially prioritize the material science aspect over the social science aspect of experimental archaeology.
+This research orientation prioritizes the material science aspect over the social science aspect of experimental archaeology.
 Similarly, standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023].
@@
 
-On the other hand, less-controlled experiments, which have been traditionally known as naturalistic or actualistic expeirments [see terminological critiques of @eren2024; @conrad2023], pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
+On the other hand, less-controlled experiments, which have been traditionally known as naturalistic or actualistic experiments [see @conrad2023; @eren2024 for detailed terminological critiques], pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008].
 Back to the cases of cultural transmission experiments, a less-controlled experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals.
-Unlike strictly controlled experiments testing one variable of interest each time, variation could be easily observed in less-controlled experiments by design.
-This feature is crucial and cannot be simply replaced by ethnographic records, because many palaeolithic technological components do not have analogues in contemporary non-industrial societies [e.g., @stout2002; @arthur2018].
+Unlike strictly controlled experiments testing one variable of interest each time [@almaatouq2024], less-controlled experiments are designed to produce variation and their interactions.
+This feature is crucial and cannot be simply replaced by ethnographic records or ethnoarchaeology, because many paleolithic technological components do not have analogs in contemporary non-industrial societies [e.g., @arthur2018; @stout2002].
 While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by less-controlled experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
-Naturalistic experiment can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
-[-@lin2018: 680-681], who proposed that the interaction between less-controlled and strictly controlled experiment "operates in a cyclical form of induction and deduction."
+Less-controlled experiments can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
+[-@lin2018: 680-681], who proposed that the interaction between less-controlled and strictly controlled experiments "operates in a cyclical form of induction and deduction."
 
@@
 
-Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker2004], tend to be restrained by the cognitive bias known as the 'curse of knowledge' or 'curse of expertise'.
-This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999], but it has further implications for the sample representativeness in experimental archaeology.
+Traditional practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with the dual identity of also being a researcher [@whittaker2004], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise." This psychological term originally refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices, when communicating with others [@hinds1999], but it has further implications for the sample representativeness in experimental archaeology.
 When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654].
 The existence of this cognitive bias is not inherently bad, and these many years of experience should be appreciated and celebrated by experimental archaeologists.
-However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a vast range of technological diversity.
+However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, have been constantly framed as generalizations regarding the evolution of technology and cognition that masks a vast range of technological diversity.
 
@@
 It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] ethnography.
-Likewise, novices' lack of expertise also helps to mitigate the 'curse of knowledge' bias that may hinder expert knappers.
+Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
 Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
 
-Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artefacts.
+Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artifacts.
 Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
@@
 Equifinality refers to situations in which a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
-While we cannot fully solve these two problems and accurately reconstruct the past behavioural processes simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation and draw a more informed inference [@reynolds1999].
+While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation and draw a more informed inference [@reynolds1999].
 The importance of specifying and documenting the context information of both the experiment and the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].
@@
 Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials.
-It can exist in the form of spreadsheets containing detailed technological attributes, photos and illustrations, or high-resolution 3D scans of individual artefacts or a whole assemblage.
+It can exist in the form of spreadsheets containing detailed technological attributes, photos and illustrations, or high-resolution 3D scans of individual artifacts or a whole assemblage.
 No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data.
@@
 
-While systematic behavioural coding methods widely used in the study of non-human animal behaviour [@fragaszy2018] are still largely neglected among archaeologists, attempts to reconstruct behavioural sequences involved in the manufacture of material remains are not infrequent.
+While systematic behavioral coding methods widely used in the study of non-human animal behavior [@fragaszy2018] are still largely neglected among archaeologists, attempts to reconstruct behavioral sequences involved in the manufacture of material remains are not infrequent.
 One such example is the cognigram, which was first systematically developed and applied in archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014; @haidle2023; @haidle2010].
-A cognigram is a graphical representation of the reconstructed behaviour behind archaeological artefacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
+A cognigram is a graphical representation of the reconstructed behavior behind archaeological artifacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
 This approach provides an elegant descriptive methodology yet is limited by its normative and analytical orientation, meaning it cannot handle variation very well.
-To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and reflects the analyst's own causal perception.
+To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and reflects the analyst's own causal understanding.
 However, this may be biased because 1) certain causal insights in stone fracture mechanics remain opaque to academic knappers until they are revealed through controlled experiments by Dibble and his colleagues [@li2022] 2) ethnographic studies demonstrated that expert non-academic practitioners can have a different set of causal understanding [@harris2021].
@@
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
-Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021; @cristino2010], etc.
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of technological complexity [@stout2021], or behavioral similarity across individuals [@mobbs2021; @cristino2010], etc.
 
@@
 df1 <- data.frame("Action" = c("Percussion", "Light Percussion", "Core Move", "Core Reposition", "Grinding", "Tool Change", "Winding Up"), 
-                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation", "Movement of the core without a change in grip, which often occurs during the core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand, which is often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone.
+                 "Definition" = c("Forcefully striking core with percussor (hammerstone or antler billet) in such a way as to potentially remove a flake.
+Each strike should be counted as a single action.", "Lower amplitude tangential strike to the tool edge of the kind often employed for platform preparation.
+Each strike should be counted as a single action.", "Movement of the core without a change in grip, which often occurs during the core inspection", "Rotation and/or reorientation of the core involving repositioning of the hand, which is often associated with the transition to a new percussion target", "Abrasion of core edges using a hammerstone.
 The abrasion movement should come from at least two different directions.", "Exchange of one percussor for another", "Preparational percussor movements towards the core that do not lead to the detachment of flakes, which can either be in direct contact with cores or not."), check.names=FALSE)
@@
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
-One potential application of ethnographic methods in experimental archaeology of stone artefacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
-Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely 'an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen).' In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography become necessary.
+One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely "an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen)." In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography becomes necessary.
 
@@
 As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project.
-Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researcher; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
+Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researchers; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
 
@@
 Through the broadening of traditional data types and recording methods revolving around experimental replicas *per se*, the Triple P conceptual framework allows the amplified multiscale expression of material cultural variation.
-It is also compatible with many theoretical orientations, ranging from behavioural archaeology (emphasis on video recording of behavioural processes) through evolutionary archaeology (emphasis on the amplification of variation) to post-processual archaeology (emphasis on perception through ethnography).
+It is also compatible with many theoretical orientations, ranging from behavioral archaeology (emphasis on video recording of behavioral processes) through evolutionary archaeology (emphasis on the amplification of variation) to post-processual archaeology (emphasis on perception through ethnography).
 In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.
-The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework has enormous potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
+The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework have enormous potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
 
-# Acknowledgements
+# Acknowledgments
 
-I thank Dietrich Stout for his helpful comments on earlier drafts of this article and Mark Moore for an inspiring discussion.
-This study was supported by the Leakey Foundation research grant titled 'Inferring skill reproduction from stone artifacts: A middle‐range approach' as well as the International Society of Human Ethology's Owen Aldis Award project titled 'Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission.'
+I thank Dietrich Stout for his helpful comments on earlier drafts of this article as well as Mark Moore and Shaya Jannati for inspiring discussions.
+This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."
 

### 2024-02-14 manuscript/manuscript_word.Rmd
revision following Dietz's advice and submitted to advances in archaeological practice.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
-abstract: "Abstract: This paper presents the Perception-Process-Product (‘Triple P’) framework that aims to expand the scope of experimental archaeology.
+abstract: "Abstract: This paper presents the Perception-Process-Product (\"Triple P\") framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.

### 2024-05-13 manuscript/manuscript_new.qmd
1st round revision submitted to Advances in Archaeological Practice.

--- 2024-02-14 manuscript/manuscript_word.Rmd
@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
-abstract: "Abstract: This paper presents the Perception-Process-Product (\"Triple P\") framework that aims to expand the scope of experimental archaeology.
+abstract: "This paper presents the Perception-Process-Product (\"Triple P\") framework that aims to expand the scope of experimental archaeology.
 The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
 It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
-\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"
+Serving as a heuristic device, this alternative mode of knowledge production is highly flexible in nature, where each single component is detachable as dicated by individual research question."
 date: "`r Sys.Date()`"
-output:
-  word_document
+fontsize: 11pt
+format: pdf
+header-includes:
+   - \usepackage{textcomp}
+   - \usepackage{utopia}
+   - \usepackage{lineno}
+   - \usepackage[utf8]{inputenc}
+   - \linenumbers
+   - \usepackage{float}
+   - \floatplacement{figure}{H}
+   - \usepackage{caption}
+   - \captionsetup[figure]{font=scriptsize}
+   - \captionsetup[table]{font=scriptsize}
 linestretch: 1.5
-link-citations: yes
+link-citations: true
 linkcolor: blue
@@
 csl: apa.csl
-fontsize: 11pt
-always_allow_html: true
 ---
@@
 Over-imitation means the copying of actions that are causally irrelevant in a goal-directed action sequence [@lyons2007].
-It is a psychological propensity that was suggested to be uniquely prevalent among humans in inter-species [@horner2005; @clay2018] comparisons and cross-cultural contexts [@nielsen2010; @stengelin2020].
+It is a psychological propensity that was suggested to be uniquely prevalent among humans when comparing with non-human primates including chimpanzees [@horner2005], bonobos [@clay2018] and orangutans [@nielsenFailureFindOverimitation2010].
+Subsequent research further suggested that within human societies over-imitation has been commonly observed among children across various cultural contexts [@nielsen2010; @stengelin2020; @nielsen2014; @subiaul2016; @berl2015].
 Gergely and Csisbra [-@gergely2006] introduced "Sylvia's Recipe" that vividly illustrates this cognitive process in the transmission of technical skills.
@@
 
-Built upon this critique of *Homo economicus* as well as intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+Built upon this critique of *Homo economicus* as well as the four strategies\^[Strategy 2 is particularly relevant here, which is "the pursuit of general questions in present-day material culture in order to acquiring laws for making useful behavioral inferences" (Schiffer, 2010: 6)] of behavioral archaeology [@schiffer2010; @reid1975], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
@@
 Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
-In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024].
+In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024; @marreiros2020].
 In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
@@
 While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by less-controlled experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023].
-Less-controlled experiments can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
+Despite the fact that one should not interpret any experiment as a direct representation of an actual past event [@eren2024], less-controlled experiments can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al.
 [-@lin2018: 680-681], who proposed that the interaction between less-controlled and strictly controlled experiments "operates in a cyclical form of induction and deduction."
@@
 It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] ethnography.
-Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Likewise, novices' lack of flintknapping expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
 Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
+Several researchers have also pointed out that the literature-informed archaeologists sometimes get lost in reconstructing previous archaeologists’ reconstructions instead of searching for diverse solutions to better understand the actual archaeological phenomenon [@bell2014; @currie2022], which is another reason why we need the presence of hobbyists and novices in the community of experimental archaeology.
 
 Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artifacts.
-Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage [@arthur2024].
 This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022; @milks2023] across different regions.
@@
 This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
-To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations [@ouzman2023].
+Moreover, considering the broader social context of the checkered disciplinary history of archaeology/anthropology featuring colonial exploitation, the changes in the evaluation system alone are not enough.
+This further highlights the need for adopting a community-based approach to fundamentally transform the power dynamics in archaeological knowledge production and distribution [@schneider2020; @lasalle2010; @atalay2012].
 
@@
 
-The implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010; @hiscock2004], two daunting challenges in archaeological inference.
+The implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@premoEquifinalityExplanationThoughts2010; @nami2010; @hiscock2004; @erenExperimentalBisonButchery2024], two daunting challenges in archaeological inference.
 Equifinality refers to situations in which a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends.
@@
 
-While systematic behavioral coding methods widely used in the study of non-human animal behavior [@fragaszy2018] are still largely neglected among archaeologists, attempts to reconstruct behavioral sequences involved in the manufacture of material remains are not infrequent.
-One such example is the cognigram, which was first systematically developed and applied in archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014; @haidle2023; @haidle2010].
+While systematic behavioral coding methods widely used in the study of non-human animal behavior [@fragaszy2018] are still largely neglected among archaeologists, attempts to reconstruct behavioral sequences involved in the manufacture of material remains are not infrequent, ranging from the well-established chaîne opératoire approach [@delage2017; @audouze2017; @soressi2011; @porqueddu2023; @dobres1999] to the more recent cognigram method.
+To illustrate the benefits and drawbacks of existing analytical frameworks, here I will use the cognigram as an example, which was first systematically developed and applied in archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2014; @haidle2023; @haidle2010].
 A cognigram is a graphical representation of the reconstructed behavior behind archaeological artifacts in chronological order of appearance [@haidle2014], which essentially represents an abstracting process of a series of action sequences achieving a similar goal.
@@
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
-Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of technological complexity [@stout2021], or behavioral similarity across individuals [@mobbs2021; @cristino2010], etc.
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of complexity of various technological systems, a classical topic in paleolithic archaeology and the evolution of human cognition [@perreault2013; @muller2017].
+Unlike the traditional approaches resorting to the extraction and comparison of lithic attributes, Stout et al.
+[-@stout2021] recorded the videos of expert flintknappers reproducing Oldowan and Acheulean technologies and then manually parsed their knapping activities using action grammar, generating multiple sequences of actions.
+Borrowing tools from computational linguistics, they then calculated the transition probability between each action category across two technological systems, which provided an objective and quantifiable proxy of measuring technological complexity.
+Another scenario of its application is the measurement of behavioral similarity across individuals [@mobbs2021; @cristino2010], which is particularly relevant in the above-mentioned cultural transmission experiments.
+Again, since the existing works in this topic mainly focuses on the sole analysis of experimental replicas, many aspects of knapping skill learning processes remain unclear.
+For example, how do different individual learning strategies (high-fidelity action copying vs.
+predominantly trial-and-error learning) affect the morphological variation of their final products?
+Or will learning behavioral conformity within a community of practice necessarily lead to the homogeneity in the formation of lithic assemblages?
+To answer these questions, the quantitative analysis of process-level data associated with the product-level data become necessary.
+Behatrix (<https://www.boris.unito.it/behatrix/>), a sister software of BORIS, allows us to calculate the action sequence (dis)similarity between novice learners and expert demonstrators/fellow novice learners using established algorithms [see @cordoni2022 for an application of analyzing play behavior similarity among gorillas].
 
@@
 hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel.
-However, ethnography has never been formally recognized as a legitimate research method in experimental archaeology.
+However, ethnography has never been formally recognized as a legitimate research method in mainstream experimental archaeology.
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
@@
 
-Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
+Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences [@day2013; @oneill2019; @therout2019], social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
 One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+For example, serial formation of step-fracture on the debitage surface are commonly interpreted as unintentional mistakes indicative of novice knappers, while in some cases researchers treat it as evidence of deliberate core rejuvenation [@akerman1993: 126].
+The accumulation of testimonies by participating knappers in terms of their intended outcome become useful in this scenario, although these materials should be examined in combination with the relevant product- and process-level data in a careful manner.
 Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely "an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen)." In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography becomes necessary.
 
-## Multi-level data curation
+## Multi-level sample and data curation
 
@@
 
+On top of the digital data curation, an easily ignored but crucial aspect regarding the integrity and reliability of research in experimental archaeology is the long-term and proper curation of psychical specimens produced during experiments, which is particularly relevant to the product-level data.
+Haythron et al.
+[-@haythorn2018] sharply pointed out that re-running statistical analyses on a publicly available spreadsheet containing incorrect lithic projectile attribute data would be meaningless.
+In this case, reexamination of the actual experiment samples become necessary.
+Moreover, it is likely that new research questions can only be answered through direct observation and measurement of the experimental assemblages themselves but not available in previously recorded data sets [@eren2016].
+The existence of these possibilities thus require that an experimental assemblage of interest should be ideally curated in public institutions with easy access, granting that their contextual information is still properly preserved [@haythorn2018].
+
 # Conclusion
@@
 In terms of its research practice, it embraces a collaborative mode of knowledge production by involving a more diverse pool of stakeholders.
-The innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework have enormous potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
+It should be noted that this alternative mode of knowledge production is not necessarily a bundle sale, where each single component is independent and detachable according to the individual research question.
+Instead, it can serve as a heuristic tool to inspire potential readers to explore a broader range of data collection and analysis strategies.
+To summarize, the innovativeness, flexibility, and inclusiveness of the Triple P conceptual framework have enormous potential in redefining what can be and what should be studied by experimental archaeology as a field and thereby contributing to a better understanding of our deep past.
 
@@
 I thank Dietrich Stout for his helpful comments on earlier drafts of this article as well as Mark Moore and Shaya Jannati for inspiring discussions.
+I am also grateful for Metin Eren and two other anonymous reviewers for their insightful feedback.
 This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."

### 2024-05-13 manuscript/manuscript_new.tex
1st round revision submitted to Advances in Archaeological Practice.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{textcomp}
\usepackage{utopia}
\usepackage{lineno}
\usepackage[utf8]{inputenc}
\linenumbers
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{caption}
\captionsetup[figure]{font=scriptsize}
\captionsetup[table]{font=scriptsize}

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework},
  pdfauthor={Cheng Liu},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Variation matters: Expanding the scope of experimental
archaeology using the Perception-Process-Product conceptual framework}
\author{Cheng Liu\footnote{Department of Anthropology, Emory University,
  Atlanta, GA, USA; cheng.liu@emory.edu}}
\date{2024-05-12}

\begin{document}
\maketitle
\begin{abstract}
This paper presents the Perception-Process-Product (``Triple P'')
framework that aims to expand the scope of experimental archaeology. The
Triple P framework emphasizes multi-level variation and interactions
across the levels of perception, process, and product to provide a more
grounded and richer explanation of the past archaeological record. It
consists of three principles: 1) acknowledging the inherent trade-off
between control and generalizability in the experimental research
design; 2) encouraging collaborative projects that involve
geographically diverse and non-traditional research participants such as
hobbyists and novices; 3) adopting a workflow that normalizes the
collection and curation of ethological and ethnographic data in
experimental projects. Serving as a heuristic device, this alternative
mode of knowledge production is highly flexible in nature, where each
single component is detachable as dicated by individual research
question.
\end{abstract}

\setstretch{1.5}
\keywords{Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production}

\section{Introduction}\label{introduction}

This paper presents the Perception-Process-Product (hereafter ``Triple
P'') conceptual framework to expand the scope of experimental
archaeology. The field has long tended to adopt the principle of Occam's
razor (e.g., \citeproc{ref-blessing2021}{Blessing \& Schmidt, 2021};
\citeproc{ref-domuxednguez-rodrigo2008}{Domínguez-Rodrigo, 2008};
\citeproc{ref-reeves2009}{Reeves et al., 2009};
\citeproc{ref-schmidt2019}{P. Schmidt et al., 2019}), whether explicitly
or implicitly. This assumption acts to center inquiry around the reverse
engineering of a past technology in a minimal or least-effort manner
while ignoring the rich contextual information experimentation affords.
When applied to the experimental study of ancient craftsmanship, Occam's
razor, or the law of parsimony, implies that a technological solution
that is simpler to reproduce is more likely to be the one used in the
archaeological context. This is insufficient to infer the preferences of
``irrational'' agents possessing incomplete information
(\citeproc{ref-armstrong2018}{Mindermann \& Armstrong, 2018}) in tool
design and use. The two conditions described here provide a better
approximation of past humans displaying extensive cultural variation as
opposed to the assumption of omniscient \emph{Homo economicus} (i.e.,
the idea that humans are consistently rational and narrowly
self-interested agents pursuing optimality) that has been rejected by
many anthropologists (\citeproc{ref-apicella2020}{Apicella et al.,
2020}; \citeproc{ref-henrichSearchHomoEconomicus2001}{Henrich et al.,
2001}). Heyes (\citeproc{ref-heyes2012}{Heyes, 2012}) similarly
questioned the abuse of parsimony in animal behavioral research and
proposed that new observational and experimental studies that allow
differential predictions to be tested become necessary when both a
simple and a complex mechanism can explain the phenomenon of interest.

In fact, there are several reasons why past technologies may violate
``parsimonious'' assumptions of minimal manufacture complexity and
optimal functional efficiency. In the evolution of technology, it is
rather common that opaque causal perception and its resulting tendency
of over-imitation can lead to the widespread and long-lasting
reproduction of technological solutions that are neither minimal in
manufacture complexity nor optimal in functional efficiency.
Over-imitation means the copying of actions that are causally irrelevant
in a goal-directed action sequence (\citeproc{ref-lyons2007}{Lyons et
al., 2007}). It is a psychological propensity that was suggested to be
uniquely prevalent among humans when comparing with non-human primates
including chimpanzees (\citeproc{ref-horner2005}{Horner \& Whiten,
2005}), bonobos (\citeproc{ref-clay2018}{Clay \& Tennie, 2018}) and
orangutans (\citeproc{ref-nielsenFailureFindOverimitation2010}{Nielsen
\& Susianto, 2010}). Subsequent research further suggested that within
human societies over-imitation has been commonly observed among children
across various cultural contexts (\citeproc{ref-berl2015}{Berl \&
Hewlett, 2015}; \citeproc{ref-nielsen2014}{Nielsen et al., 2014};
\citeproc{ref-nielsen2010}{Nielsen \& Tomaselli, 2010};
\citeproc{ref-stengelin2020}{Stengelin et al., 2020};
\citeproc{ref-subiaul2016}{Subiaul et al., 2016}). Gergely and Csisbra
(\citeproc{ref-gergely2006}{2006}) introduced ``Sylvia's Recipe'' that
vividly illustrates this cognitive process in the transmission of
technical skills. Sylvia is an education researcher who developed a
unique way of cooking ham roast by observing her mother during
childhood, where she cut both ends of a ham. Later in life, her mother
happened to watch her cooking, where she noticed and questioned the
purpose of this step of preparation. Sylvia could not answer it and was
then told that it was processed that way because her mother did not have
a pan that was large enough to cook a full-sized ham. The commonality of
this opaque causal perception has also been demonstrated in a recent
study of Hadza bowmakers. Harris et al.
(\citeproc{ref-harris2021}{2021}) found that even experienced bowmakers
only possess limited causal knowledge regarding the design and
construction of bows according to modern engineering principles, meaning
they cannot spell out the mechanical (dis)advantages of many
morphological features.

On the other hand, path dependence also constrained the pursuit of
functional optimization or simplification of manufacturing procedures.
In this case, people are implicitly or explicitly aware of the existence
of a more efficient solution but still stick to the older one due to the
cost of learning, cultural conservatism
(\citeproc{ref-acerbi2009}{Acerbi et al., 2009};
\citeproc{ref-ghirlanda2006}{Ghirlanda et al., 2006};
\citeproc{ref-morin2022}{Morin, 2022}), or other reasons. One such
example in the evolution of technology is the longevity of QWERTY
keyboard design (\citeproc{ref-kafaee2022}{Kafaee et al., 2022}). This
deliberately unergonomic solution was invented in the era of typewriters
in order to disperse commonly used letters, preventing the most
frequently struck ``hammers'' from clashing. Yet it is still the most
common keyboard design today when such constraint does not exist anymore
on modern computer hardware. In short, we should acknowledge the
existence and variation of many ``good-enough'' technological solutions
featuring various degrees of ``redundancy'' in real-world contexts,
which often represent locally adaptive peaks instead of a global optimum
in a multimodal fitness landscape due to multiple constraints and
trade-off factors (\citeproc{ref-bettinger1982}{Bettinger \& Baumhoff,
1982}; \citeproc{ref-mesoudi2008}{Mesoudi \& O'Brien, 2008}).

Built upon this critique of \emph{Homo economicus} as well as the four
strategies\^{}{[}Strategy 2 is particularly relevant here, which is
``the pursuit of general questions in present-day material culture in
order to acquiring laws for making useful behavioral inferences''
(Schiffer, 2010: 6){]} of behavioral archaeology
(\citeproc{ref-reid1975}{Reid et al., 1975};
\citeproc{ref-schiffer2010}{Schiffer, 2010}), here I propose the Triple
P framework, which aims to \textbf{a)} amplify the expression of
variation in experimental replicas (product) and their associated
behavioral channels (process) as well as sensory experiences
(perception) by experiments in diverse contexts and \textbf{b)} better
identify the complex interacting relationships across these three levels
of variations in real-world conditions. To accomplish these two
objectives, I advocate the following three principles as integral
components of the Triple P framework, which requires \textbf{1)}
acknowledging the inherent trade-off between control and
generalizability in the experimental research design and \textbf{2)}
encouraging collaborative projects that involve geographically diverse
and non-traditional research participants such as hobbyists and novices.
These two principles are developed to advocate a pluralistic approach to
the explanation of complex variation, which has received more attention
from evolutionary anthropology (\citeproc{ref-antuxf3n2017}{Antón \&
Kuzawa, 2017}) to cognitive science (\citeproc{ref-barrett2020}{Barrett,
2020}), instead of treating the optimization-based research agenda as a
panacea. The second principle particularly allows researchers to develop
research questions that are also meaningful to descendant communities
through respectful conversation and collaboration
(\citeproc{ref-montgomery2023}{Montgomery \& Fryer, 2023}). The Triple P
framework also \textbf{3)} adopts a workflow that normalizes the
collection and curation of ethological and ethnographic data in
experimental projects. It is acknowledged that strategies of data
collection and analysis of a given experimental project should be
primarily derived from the research question, but the awareness of the
rich toolkit available can sometimes inspire researchers to ask
questions that are bold and transformative
(\citeproc{ref-schmidt2020}{S. C. Schmidt \& Marwick, 2020}). Here I
will leverage the extensive corpus in experimental designs and
inferences revolving around stone artifacts to clarify its meaning and
demonstrate the necessity and potential of this framework.

\section{What good is less-controlled
experimentation?}\label{what-good-is-less-controlled-experimentation}

The trade-off between causal inference (aka ``internal validity'') and
generalization (aka ``external validity'') forms a central issue in
experimental design across different disciplines
(\citeproc{ref-eren2016}{Eren et al., 2016}; \citeproc{ref-roe2009}{Roe
\& Just, 2009}: 1266-1267). Even in fields known for their development
of rigorous and well-controlled experimental methods such as cognitive
psychology and neuroscience, researchers have started to use relatively
naturalistic stimuli more frequently and advocate a paradigm shift to
semi-controlled experiments due to the generalizability crisis, namely
the prevailing mismatch between phenomenon of interest and measured
variables in psychological science (\citeproc{ref-nastase2020}{Nastase
et al., 2020}; \citeproc{ref-shamay-tsoory2019}{Shamay-Tsoory \&
Mendelsohn, 2019}; \citeproc{ref-sonkusare2019}{Sonkusare et al., 2019};
\citeproc{ref-yarkoni2022}{Yarkoni, 2022}). In contrast, the past
decades have witnessed experimental archaeology's growing research
interests focusing on the robust inference of causal mechanisms while
compromising generalizability in the explanation of material culture
variation (\citeproc{ref-eren2016}{Eren et al., 2016};
\citeproc{ref-eren2024}{Eren \& Meltzer, 2024};
\citeproc{ref-lin2018}{Lin et al., 2018};
\citeproc{ref-marreiros2020}{Marreiros et al., 2020}). In the context of
stone artifact replication, one typical research design emphasizing
causality over generalizability is the use of knapping machines/robots
(\citeproc{ref-li2022}{Li et al., 2022};
\citeproc{ref-pfleging2019}{Pfleging et al., 2019}), which has helped
map out the physical constraints of stone artifact manufacture and use
through the identification of causal relationships between input (force,
exterior platform angle, platform depth, etc.) and outcome variables
(flake size, flake shape, wear formation, etc.). All variables of
interest in this setting are relatively easy to measure, quantify, and
control, but this type of design can be insufficient in inferring how
context-generic principles interact in a particular context as reflected
in real-world conditions. This research orientation prioritizes the
material science aspect over the social science aspect of experimental
archaeology. Similarly, standardized artificial materials like bricks
(\citeproc{ref-lombao2017}{Lombao et al., 2017}) or foam blocks
(\citeproc{ref-schillinger2016}{Schillinger et al., 2016}) have been
used to standardize materials and/or reduce learning demands in
experimental studies focusing on the transmission of lithic
technologies, with implications for the generalizability of results
(\citeproc{ref-liu2023}{Liu et al., 2023}). In real-world knapping, each
rock has a different shape and often different physical properties such
as inner cracks and inclusions, and this heterogeneity itself represents
a critical variable in cultural transmission and skill development
(\citeproc{ref-proffitt2022}{Proffitt et al., 2022}).

On the other hand, less-controlled experiments, which have been
traditionally known as naturalistic or actualistic experiments (see
\citeproc{ref-conrad2023}{Conrad et al., 2023};
\citeproc{ref-eren2024}{Eren \& Meltzer, 2024} for detailed
terminological critiques), pay more attention to how experimental
insights can be generalized to archaeological samples by incorporating
authentic materials and plausible social settings with a certain degree
of compromised control (\citeproc{ref-outram2008}{Outram, 2008}). Back
to the cases of cultural transmission experiments, a less-controlled
experiment would involve the use of natural rocks with varied morphology
instead of standardized artificial materials as well as human
demonstrators instead of videos of knapping instruction, despite the
fact that the latter will remain consistent across individuals. Unlike
strictly controlled experiments testing one variable of interest each
time (\citeproc{ref-almaatouq2024}{Almaatouq et al., 2024}),
less-controlled experiments are designed to produce variation and their
interactions. This feature is crucial and cannot be simply replaced by
ethnographic records or ethnoarchaeology, because many paleolithic
technological components do not have analogs in contemporary
non-industrial societies (e.g., \citeproc{ref-arthur2018}{Arthur, 2018};
\citeproc{ref-stout2002}{Stout, 2002}). While uncontrolled variation has
traditionally been viewed as highly problematic, statistical techniques
for developing causal inference from observational data, of the kind
produced by less-controlled experiments, have also been greatly boosted
in epidemiology and economics in recent years
(\citeproc{ref-cunningham2021}{Cunningham, 2021};
\citeproc{ref-hernan2023}{Hernan \& Robins, 2023}). Despite the fact
that one should not interpret any experiment as a direct representation
of an actual past event (\citeproc{ref-eren2024}{Eren \& Meltzer,
2024}), less-controlled experiments can serve a heuristic role in
hypothesis generation, aligning with the perspective of Lin et al.
(\citeproc{ref-lin2018}{2018}: 680-681), who proposed that the
interaction between less-controlled and strictly controlled experiments
``operates in a cyclical form of induction and deduction.''

\section{Many places, many voices}\label{many-places-many-voices}

Traditional practices in experimental archaeology, as manifested by the
fact that a majority of scholarly publications are produced as results
of experiments conducted by a single knapper with the dual identity of
also being a researcher (\citeproc{ref-whittaker2004}{Whittaker, 2004}),
tend to be restrained by the cognitive bias known as the ``curse of
knowledge'' or ``curse of expertise.'' This psychological term
originally refers to the phenomenon that it is extremely challenging for
experts to ignore the information that is held by them but not others,
particularly novices, when communicating with others
(\citeproc{ref-hinds1999}{Hinds, 1999}), but it has further implications
for the sample representativeness in experimental archaeology. When the
knapping expertise is gradually formed through multiple years of
observations and trial-and-error learning, an expert knapper develops
some specific ways of strategic planning, motor habits (and their
associated impacts on anatomical forms like wrist and elbow),
preferences of percussor and raw material types, as well as familiarity
of various techniques that become unforgettable
(\citeproc{ref-moore2020}{Moore, 2020}: 654). The existence of this
cognitive bias is not inherently bad, and these many years of experience
should be appreciated and celebrated by experimental archaeologists.
However, what is problematic is that the results of replication
experiments conducted by these experienced practitioners, often in
settings of single knapper, have been constantly framed as
generalizations regarding the evolution of technology and cognition that
masks a vast range of technological diversity.

Modern flintknapping techniques, as a research subject and a scientific
method, originated from hobbyists' individualistic trials of reverse
engineering during the 19th century (\citeproc{ref-coles1979}{Coles,
1979}; \citeproc{ref-flenniken1984}{Flenniken, 1984};
\citeproc{ref-johnson1978}{Johnson, 1978};
\citeproc{ref-whittaker1994}{Whittaker, 1994}: 54-61). Hobbyist knappers
represent a huge repertoire of technological knowledge that does not
fully overlap with what is acquired by academic knappers. They tend to
generate ideas that may appear to be counter-intuitive at first glance
for academics. One such example is the utility of obtuse edge angle as
demonstrated by Don Crabtree (\citeproc{ref-crabtree1977}{1977}), a
mostly self-educated flintknapper yet one of the most important figures
in experimental archaeology. In his experiment, Crabtree demonstrated
the excellent performance of blade dorsal ridge on tasks like shaving
and cutting hard materials, challenging the traditional perspective on
producing sharp lateral edges as the sole purpose of stone toolmaking
and shedding light on future functional reconstruction through the
use-wear analysis. It is rather unfortunate that collaborations between
academics and hobbyists are less common than expected due to their
complicated and uneasy relationships as detailed in Whittaker's
(\citeproc{ref-whittaker2004}{2004}) ethnography. Likewise, novices'
lack of flintknapping expertise also helps to mitigate the ``curse of
knowledge'' bias that may hinder expert knappers. Their involvement can
potentially lead to the discovery of alternative methods, techniques,
and interpretations that may have been overlooked by experts. Several
researchers have also pointed out that the literature-informed
archaeologists sometimes get lost in reconstructing previous
archaeologists' reconstructions instead of searching for diverse
solutions to better understand the actual archaeological phenomenon
(\citeproc{ref-bell2014}{Bell, 2014}; \citeproc{ref-currie2022}{Currie,
2022}), which is another reason why we need the presence of hobbyists
and novices in the community of experimental archaeology.

Emphasizing variation at its core, the Triple P conceptual framework
recognizes that experimental archaeology can greatly benefit from
diverse perspectives (\citeproc{ref-pargeter2023}{Pargeter et al.,
2023}: 164) and thereby inherently adopts a collaborative mode of
knowledge production, which has been recently advocated in experimental
studies (\citeproc{ref-liuInferringCulturalReproduction2023}{Liu \&
Stout, 2023}; \citeproc{ref-ranhorn2020}{Ranhorn et al., 2020}) and
museum collection studies (\citeproc{ref-timbrell2023}{Timbrell, 2023})
of stone artifacts. Furthermore, the Triple P framework acknowledges
that communities living in specific geographical areas possess unique
insights and understanding of their cultural heritage
(\citeproc{ref-arthur2024}{Arthur et al., 2024}). This emphasis on team
efforts and inclusivity allows for a more complete understanding of the
non-utilitarian or unexpected aspects of raw material procurement
(\citeproc{ref-batalla2016}{Batalla, 2016}) and selection
(\citeproc{ref-arthur2021}{Arthur, 2021}), pre-treatment
(\citeproc{ref-maloney2020}{Maloney \& Street, 2020}), production
(\citeproc{ref-griffin2013}{Griffin et al., 2013}), and use
(\citeproc{ref-martellotta2022}{Martellotta et al., 2022};
\citeproc{ref-milks2023}{Milks et al., 2023}) across different regions.
Through ethical collaborations with those knapping practitioners in
non-industrial societies in the research process, the framework allows
their voices to be heard and their contributions to be acknowledged.
This not only enhances the quality of research outcomes but also fosters
a sense of ownership and pride within these communities, strengthening
the connection between archaeological research and the people it
directly affects (\citeproc{ref-douglass2020}{Douglass, 2020};
\citeproc{ref-marshall2002}{Marshall, 2002};
\citeproc{ref-montgomery2023}{Montgomery \& Fryer, 2023}).

However, the facilitation of large-scale collaborations faces challenges
within the current system of research evaluation. The prevailing
practice of attributing credit primarily to the first author and senior
(last/corresponding) author in peer-reviewed journal papers hampers the
recognition of multiple contributors. This system often overlooks the
valuable input of collaborators who may not fit into the traditional
authorship structure but have made significant intellectual and
practical contributions to the research. To truly embrace the principles
of collaboration and inclusivity, there is a need for a reevaluation of
the research evaluation system, allowing for proper acknowledgment of
the diverse voices and contributions involved in large-scale
collaborations (\citeproc{ref-ouzman2023}{Ouzman, 2023}). Moreover,
considering the broader social context of the checkered disciplinary
history of archaeology/anthropology featuring colonial exploitation, the
changes in the evaluation system alone are not enough. This further
highlights the need for adopting a community-based approach to
fundamentally transform the power dynamics in archaeological knowledge
production and distribution (\citeproc{ref-atalay2012}{Atalay, 2012};
\citeproc{ref-lasalle2010}{La Salle, 2010};
\citeproc{ref-schneider2020}{Schneider \& Hayes, 2020}).

\section{The Triple P framework in
action}\label{the-triple-p-framework-in-action}

The implementation of the Triple P framework involves the collection of
process-level (ethological) and perception-level (ethnographic) data
(\textbf{Figure} @ref(fig:concept)), which is critical to address
equifinality and multifinality
(\citeproc{ref-erenExperimentalBisonButchery2024}{Eren et al., 2024};
\citeproc{ref-hiscock2004}{Hiscock, 2004}; \citeproc{ref-nami2010}{Nami,
2010}; \citeproc{ref-premoEquifinalityExplanationThoughts2010}{Premo,
2010}), two daunting challenges in archaeological inference.
Equifinality refers to situations in which a similar state or
consequence can be achieved through multiple different paths, while
multifinality emerges when a similar process can lead to multiple ends.
While we cannot fully solve these two problems and accurately
reconstruct the past behavioral processes simply based on materials
remains, context-rich experiments involving the collection of
ethological and ethnographic data can help us better document an
enlarged range of possible combinations of variation and draw a more
informed inference (\citeproc{ref-reynolds1999}{Reynolds, 1999}). The
importance of specifying and documenting the context information of both
the experiment and the phenomenon of interest has also been recently
highlighted in psychological sciences
(\citeproc{ref-holleman2020}{Holleman et al., 2020}).

\begin{figure}[H]

{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{../figure/Fig1.jpg}

}

\caption{A schematic diagram demonstrating how to operationalize the
Perception-Process-Product conceptual framework.}

\end{figure}%

\subsection{Product-level data}\label{product-level-data}

Traditionally speaking, the product-level data, namely the documentation
and analysis of replicas, form the sole research subject of experimental
archaeology and serve as the tangible foundation for analogical
inference in the interpretation of archaeological materials. It can
exist in the form of spreadsheets containing detailed technological
attributes, photos and illustrations, or high-resolution 3D scans of
individual artifacts or a whole assemblage. No particular modification
regarding the collection procedure of product-level data is required in
the context of the Triple P framework, although the definition of
variables measured and the documentation techniques (models of
camera/scanners, light setting, processing software version and
workflow, etc.) should be always available in the relevant meta-data. I
also strongly recommend adopting good habits in spreadsheet data
organization (\citeproc{ref-broman2018}{Broman \& Woo, 2018}).

\subsection{Process-level data}\label{process-level-data}

While systematic behavioral coding methods widely used in the study of
non-human animal behavior (\citeproc{ref-fragaszy2018}{Fragaszy \&
Mangalam, 2018}) are still largely neglected among archaeologists,
attempts to reconstruct behavioral sequences involved in the manufacture
of material remains are not infrequent, ranging from the
well-established chaîne opératoire approach
(\citeproc{ref-audouze2017}{Audouze et al., 2017};
\citeproc{ref-delage2017}{Delage, 2017};
\citeproc{ref-dobres1999}{Dobres, 1999};
\citeproc{ref-porqueddu2023}{Porqueddu et al., 2023};
\citeproc{ref-soressi2011}{Soressi \& Geneste, 2011}) to the more recent
cognigram method. To illustrate the benefits and drawbacks of existing
analytical frameworks, here I will use the cognigram as an example,
which was first systematically developed and applied in archaeological
research by Haidle (\citeproc{ref-haidleHowThinkSimple2009}{Haidle,
2009}, \citeproc{ref-haidle2010}{2010}, \citeproc{ref-haidle2014}{2014},
\citeproc{ref-haidle2023}{2023}). A cognigram is a graphical
representation of the reconstructed behavior behind archaeological
artifacts in chronological order of appearance
(\citeproc{ref-haidle2014}{Haidle, 2014}), which essentially represents
an abstracting process of a series of action sequences achieving a
similar goal. This approach provides an elegant descriptive methodology
yet is limited by its normative and analytical orientation, meaning it
cannot handle variation very well. To some extent, it describes the
minimal steps to achieve a goal from the perspective of reverse
engineering and reflects the analyst's own causal understanding.
However, this may be biased because 1) certain causal insights in stone
fracture mechanics remain opaque to academic knappers until they are
revealed through controlled experiments by Dibble and his colleagues
(\citeproc{ref-li2022}{Li et al., 2022}) 2) ethnographic studies
demonstrated that expert non-academic practitioners can have a different
set of causal understanding (\citeproc{ref-harris2021}{Harris et al.,
2021}).

Consequently, we need to accumulate more real-world data by recording a
large number of toolmaking videos and conducting systematic ethogram
analysis. With the emergence of new software platforms such as BORIS
(\citeproc{ref-friard2016}{Friard \& Gamba, 2016}), the difficulty of
coding has decreased significantly in recent years (\textbf{Figure}
@ref(fig:ethogram)). Here I use a modified version of action grammar
developed by (\citeproc{ref-stout2021}{Stout et al., 2021}) as an
example, among multiple coding schemes featuring different research
focus (\citeproc{ref-muller2023}{Muller et al., 2023}) or granularity
(\citeproc{ref-cueva-temprana2019}{Cueva-Temprana et al., 2019};
\citeproc{ref-mahaney2014}{Mahaney, 2014}; \citeproc{ref-roux2005}{Roux
\& David, 2005}). The knapping action recorded in videos can be coded
following the ethogram presented in \textbf{Table} @ref(tab:tab1).
Depending on the original research question, sequences of coded actions
can then be used in further analysis, such as the measurement of
complexity of various technological systems, a classical topic in
paleolithic archaeology and the evolution of human cognition
(\citeproc{ref-muller2017}{Muller et al., 2017};
\citeproc{ref-perreault2013}{Perreault et al., 2013}). Unlike the
traditional approaches resorting to the extraction and comparison of
lithic attributes, Stout et al. (\citeproc{ref-stout2021}{2021})
recorded the videos of expert flintknappers reproducing Oldowan and
Acheulean technologies and then manually parsed their knapping
activities using action grammar, generating multiple sequences of
actions. Borrowing tools from computational linguistics, they then
calculated the transition probability between each action category
across two technological systems, which provided an objective and
quantifiable proxy of measuring technological complexity. Another
scenario of its application is the measurement of behavioral similarity
across individuals (\citeproc{ref-cristino2010}{Cristino et al., 2010};
\citeproc{ref-mobbs2021}{Mobbs et al., 2021}), which is particularly
relevant in the above-mentioned cultural transmission experiments.
Again, since the existing works in this topic mainly focuses on the sole
analysis of experimental replicas, many aspects of knapping skill
learning processes remain unclear. For example, how do different
individual learning strategies (high-fidelity action copying
vs.~predominantly trial-and-error learning) affect the morphological
variation of their final products? Or will learning behavioral
conformity within a community of practice necessarily lead to the
homogeneity in the formation of lithic assemblages? To answer these
questions, the quantitative analysis of process-level data associated
with the product-level data become necessary. Behatrix
(\url{https://www.boris.unito.it/behatrix/}), a sister software of
BORIS, allows us to calculate the action sequence (dis)similarity
between novice learners and expert demonstrators/fellow novice learners
using established algorithms (see \citeproc{ref-cordoni2022}{Cordoni et
al., 2022} for an application of analyzing play behavior similarity
among gorillas).

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{../figure/Fig2.jpg}

}

\caption{An example of coding a handaxe knapping session using the BORIS
software.}

\end{figure}%

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.0914}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.9086}}@{}}
\caption{A modified version of the original action grammar presented in
(\citeproc{ref-stout2021}{Stout et al., 2021})}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Action
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Definition
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Action
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Definition
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Percussion & Forcefully striking core with percussor (hammerstone or
antler billet) in such a way as to potentially remove a flake. Each
strike should be counted as a single action. \\
Light Percussion & Lower amplitude tangential strike to the tool edge of
the kind often employed for platform preparation. Each strike should be
counted as a single action. \\
Core Move & Movement of the core without a change in grip, which often
occurs during the core inspection \\
Core Reposition & Rotation and/or reorientation of the core involving
repositioning of the hand, which is often associated with the transition
to a new percussion target \\
Grinding & Abrasion of core edges using a hammerstone. The abrasion
movement should come from at least two different directions. \\
Tool Change & Exchange of one percussor for another \\
Winding Up & Preparational percussor movements towards the core that do
not lead to the detachment of flakes, which can either be in direct
contact with cores or not. \\
\end{longtable}

\subsection{Perception-level data}\label{perception-level-data}

Ethnographies revolving around experimental archaeology as a field
(\citeproc{ref-reevesflores2012}{Reeves Flores, 2012}), as well as
practices of specific technologies like flintknapping, including
contemporary U.S. hobbyists (\citeproc{ref-whittaker2004}{Whittaker,
2004}) and knapping practitioners in various non-industrial societies
(\citeproc{ref-arthur2018}{Arthur, 2018};
\citeproc{ref-stout2002}{Stout, 2002}), are far from novel. However,
ethnography has never been formally recognized as a legitimate research
method in mainstream experimental archaeology. Echoing with the recent
trends of adopting embodied cognition (\citeproc{ref-varela2017}{Varela
et al., 2017}) in archaeological research
(\citeproc{ref-malafouris2013}{Malafouris, 2013}), ethnographic data and
methods can reveal hidden information (e.g., intention, phenomenology)
that is otherwise irretrievable and thus should occupy a unique niche in
experimental archaeology. Within the broader context of burgeoning
interest in mixed-method research in contemporary social science
(\citeproc{ref-creswell2017}{Creswell \& Clark, 2017}), this also echoes
the post-positivist turn in psychology in the past decades, particularly
the emphasis on the value of incorporating qualitative research
(\citeproc{ref-stoutCognitiveScienceTechnology2021}{Stout, 2021};
\citeproc{ref-syed2022}{Syed \& McLean, 2022};
\citeproc{ref-weger2019}{Weger et al., 2019}).

Through participant observation, interviews, and detailed field notes,
ethnography can capture the subtle nuances of perception, such as
cognitive affordances (\citeproc{ref-hussain2021}{Hussain \& Will,
2021}; \citeproc{ref-roepstorff2008}{Roepstorff, 2008}), sensory
experiences (\citeproc{ref-day2013}{Day, 2013};
\citeproc{ref-oneill2019}{O'Neill \& O'Sullivan, 2019};
\citeproc{ref-therout2019}{Skeates \& Day, 2019}), social interactions,
and cultural meanings associated with the experimental activities
(\citeproc{ref-gowlland2019}{Gowlland, 2019}). Compared with the
ethological methods, the interview questions and participant observation
in ethnographic methods feature an even higher degree of freedom and
rely more heavily on the research question as well as ad-hoc
interaction. One potential application of ethnographic methods in
experimental archaeology of stone artifacts is asking knappers about the
intentions of each action and see how it matches with the results as
revealed by lithic analysis of replicas, which can provide crucial
contextual information addressing the issues of equifinality and
multifinality in the formation of lithic assemblage. For example, serial
formation of step-fracture on the debitage surface are commonly
interpreted as unintentional mistakes indicative of novice knappers,
while in some cases researchers treat it as evidence of deliberate core
rejuvenation (\citeproc{ref-akerman1993}{Akerman, 1993}: 126). The
accumulation of testimonies by participating knappers in terms of their
intended outcome become useful in this scenario, although these
materials should be examined in combination with the relevant product-
and process-level data in a careful manner. Instead of seeing intention
as something abstruse or unapproachable in archaeology
(\citeproc{ref-david2004}{David, 2004};
\citeproc{ref-russell2004}{Russell, 2004}), the Triple P framework
adopts a novel definition proposed by Quillien and German
(\citeproc{ref-quillien2021}{2021}: 1) from the perspective of causal
perception, namely ``an agent did X \emph{intentionally} to the extent
that X was causally dependent on how much the agent wanted X to happen
(or not to happen).'' In this sense, the mismatch between how different
individuals perceive cause-and-effect relationships and how they are
organized according to physical laws is exactly where interesting
variation emerges and where ethnography becomes necessary.

\subsection{Multi-level sample and data
curation}\label{multi-level-sample-and-data-curation}

The comparative study and large-scale synthesis of variation data
require the building of centralized, open-access, and carefully curated
data infrastructure (\citeproc{ref-marwick2018}{Marwick \& Birch,
2018}), which unfortunately still does not exist yet in experimental
archaeology. The accessibility and availability of experimental data can
foster collaboration and enhance the reproducibility and transparency of
research findings, as others can verify and validate the results by
examining the original data. Moreover, a centralized database also
promotes data preservation and long-term accessibility. By storing
experimental data in a structured and organized manner, it safeguards
valuable information from potential loss or degradation over time. This
preservation ensures that the data remains accessible for future
researchers, avoiding the loss of valuable insights and preventing the
need for redundant and costly repetitions of experiments. It also allows
for the reanalysis of existing data, facilitating discoveries and
insights that may not have been initially anticipated. However, it has
been widely acknowledged that the reuse of archaeological data has not
received enough attention among researchers in our discipline
(\citeproc{ref-faniel2018}{Faniel et al., 2018};
\citeproc{ref-huggett2018}{Huggett, 2018};
\citeproc{ref-moody2021}{Moody et al., 2021}).

Among the three dimensions of the Triple P framework, the product-level
data are usually stored in the format of spreadsheets, photos, and 3D
models, and the perception-level data formats mainly include audio files
and their transcribed texts, whereas videos are the main vector of
process-level data, a rather non-traditional data format in
archaeological research featuring the largest file size compared with
the other two. As such, following data sharing principles of FAIR
(\citeproc{ref-wilkinson2016}{Wilkinson et al., 2016}) and CARE
(\citeproc{ref-carroll2020}{Carroll et al., 2020}), the Triple P
framework recommends Databrary (\citeproc{ref-gilmore2015}{Gilmore et
al., 2015}; \citeproc{ref-simon2015}{Simon et al., 2015}), a web-based
library originally designed for developmental scientists, as the main
data curation platform, where researchers can freely upload video files
with no size limit and related metadata that can connect with different
types of data within the same project. Databrary has three advantages
compared with other data storage solutions: a) no cost from the side of
researchers; 2) long-term data security monitored by a specialized
maintenance team; and 3) fostering potential collaborations between
experimental archaeologists and developmental psychologists.

On top of the digital data curation, an easily ignored but crucial
aspect regarding the integrity and reliability of research in
experimental archaeology is the long-term and proper curation of
psychical specimens produced during experiments, which is particularly
relevant to the product-level data. Haythron et al.
(\citeproc{ref-haythorn2018}{2018}) sharply pointed out that re-running
statistical analyses on a publicly available spreadsheet containing
incorrect lithic projectile attribute data would be meaningless. In this
case, reexamination of the actual experiment samples become necessary.
Moreover, it is likely that new research questions can only be answered
through direct observation and measurement of the experimental
assemblages themselves but not available in previously recorded data
sets (\citeproc{ref-eren2016}{Eren et al., 2016}). The existence of
these possibilities thus require that an experimental assemblage of
interest should be ideally curated in public institutions with easy
access, granting that their contextual information is still properly
preserved (\citeproc{ref-haythorn2018}{Haythorn et al., 2018}).

\section{Conclusion}\label{conclusion}

Through the broadening of traditional data types and recording methods
revolving around experimental replicas \emph{per se}, the Triple P
conceptual framework allows the amplified multiscale expression of
material cultural variation. It is also compatible with many theoretical
orientations, ranging from behavioral archaeology (emphasis on video
recording of behavioral processes) through evolutionary archaeology
(emphasis on the amplification of variation) to post-processual
archaeology (emphasis on perception through ethnography). In terms of
its research practice, it embraces a collaborative mode of knowledge
production by involving a more diverse pool of stakeholders. It should
be noted that this alternative mode of knowledge production is not
necessarily a bundle sale, where each single component is independent
and detachable according to the individual research question. Instead,
it can serve as a heuristic tool to inspire potential readers to explore
a broader range of data collection and analysis strategies. To
summarize, the innovativeness, flexibility, and inclusiveness of the
Triple P conceptual framework have enormous potential in redefining what
can be and what should be studied by experimental archaeology as a field
and thereby contributing to a better understanding of our deep past.

\section{Acknowledgments}\label{acknowledgments}

I thank Dietrich Stout for his helpful comments on earlier drafts of
this article as well as Mark Moore and Shaya Jannati for inspiring
discussions. I am also grateful for Metin Eren and two other anonymous
reviewers for their insightful feedback. This study was supported by the
Leakey Foundation research grant titled ``Inferring skill reproduction
from stone artifacts: A middle‐range approach'' as well as the
International Society of Human Ethology's Owen Aldis Award project
titled ``Sealed in stones: The computational ethology of stone
toolmaking and its implications to the evolution of cultural
transmission.''

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-acerbi2009}
Acerbi, A., Enquist, M., \& Ghirlanda, S. (2009). Cultural evolution and
individual development of openness and conservatism. \emph{Proceedings
of the National Academy of Sciences}, \emph{106}(45), 18931--18935.
\url{https://doi.org/10.1073/pnas.0908889106}

\bibitem[\citeproctext]{ref-akerman1993}
Akerman, K. (1993). The status of the horsehoof core. \emph{Records of
the Australian Museum, Supplement}, \emph{17}, 125127.
\url{https://media.australian.museum/media/Uploads/Journals/17795/64_complete.pdf}

\bibitem[\citeproctext]{ref-almaatouq2024}
Almaatouq, A., Griffiths, T. L., Suchow, J. W., Whiting, M. E., Evans,
J., \& Watts, D. J. (2024). Beyond playing 20 questions with nature:
Integrative experiment design in the social and behavioral sciences.
\emph{Behavioral and Brain Sciences}, \emph{47}, e33.
\url{https://doi.org/10.1017/S0140525X22002874}

\bibitem[\citeproctext]{ref-antuxf3n2017}
Antón, S. C., \& Kuzawa, C. W. (2017). Early homo, plasticity and the
extended evolutionary synthesis. \emph{Interface Focus}, \emph{7}(5),
20170004. \url{https://doi.org/10.1098/rsfs.2017.0004}

\bibitem[\citeproctext]{ref-apicella2020}
Apicella, C., Norenzayan, A., \& Henrich, J. (2020). Beyond WEIRD: A
review of the last decade and a look ahead to the global laboratory of
the future. \emph{Evolution and Human Behavior}, \emph{41}(5), 319--329.
\url{https://doi.org/10.1016/j.evolhumbehav.2020.07.015}

\bibitem[\citeproctext]{ref-arthur2018}
Arthur, K. W. (2018). \emph{The lives of stone tools: Crafting the
status, skill, and identity of flintknappers}. University of Arizona
Press.

\bibitem[\citeproctext]{ref-arthur2021}
Arthur, K. W. (2021). Material Scientists: Learning the Importance of
Colour and Brightness from Lithic Practitioners. \emph{Cambridge
Archaeological Journal}, \emph{31}(2), 293--304.
\url{https://doi.org/10.1017/S0959774320000347}

\bibitem[\citeproctext]{ref-arthur2024}
Arthur, K. W., Barkai, R., Allen, C., Shpayer, E. A., Efrati, B.,
Finkel, M., Ganchrow, D., Horowitz, R. A., Litov, V., Lombard, M.,
Sillitoe, P., \& Swenson, E. (2024). Ancestral Stones and Stone Stories:
Reimagining Human Relationships with Stone from the Paleolithic to the
Present. \emph{Archaeologies}, \emph{20}(1), 1--23.
\url{https://doi.org/10.1007/s11759-024-09502-y}

\bibitem[\citeproctext]{ref-atalay2012}
Atalay, S. (2012). \emph{Community-based archaeology: Research with, by,
and for indigenous and local communities}. University of California
Press.
\url{https://books.google.com/books?hl=en&lr=&id=q7MwDwAAQBAJ&oi=fnd&pg=PR9&dq=sonya+atalay+community+based+archaeology&ots=EoTKMq3e9j&sig=V-q0p0F6b546SQt9lhW7C59Hejw}

\bibitem[\citeproctext]{ref-audouze2017}
Audouze, F., Bodu, P., Karlin, C., Julien, M., Pelegrin, J., \& Perlès,
C. (2017). Leroi-gourhan and the chaîne opératoire: A response to
delage. \emph{World Archaeology}, \emph{49}(5), 718723.
\url{https://doi.org/10.1080/00438243.2017.1416012}

\bibitem[\citeproctext]{ref-barrett2020}
Barrett, H. C. (2020). Towards a Cognitive Science of the Human:
Cross-Cultural Approaches and Their Urgency. \emph{Trends in Cognitive
Sciences}, \emph{24}(8), 620--638.
\url{https://doi.org/10.1016/j.tics.2020.05.007}

\bibitem[\citeproctext]{ref-batalla2016}
Batalla, A. N. (2016). Studies of indigenous lithic procurement in
Uruguay and their implications for Southern Cone archaeology.
\emph{Journal of Lithic Studies}, \emph{3}(1), 265--292.
\url{https://doi.org/10.2218/jls.v3i1.1522}

\bibitem[\citeproctext]{ref-bell2014}
Bell, M. (2014). \emph{Experimental archaeology at the crossroads: A
contribution to interpretation or evidence of {`}xeroxing{'}?} (R.
Chapman \& A. Wylie, Eds.; pp. 42--58). Routledge.

\bibitem[\citeproctext]{ref-berl2015}
Berl, R. E. W., \& Hewlett, B. S. (2015). Cultural Variation in the Use
of Overimitation by the Aka and Ngandu of the Congo Basin. \emph{PLOS
ONE}, \emph{10}(3), e0120180.
\url{https://doi.org/10.1371/journal.pone.0120180}

\bibitem[\citeproctext]{ref-bettinger1982}
Bettinger, R. L., \& Baumhoff, M. A. (1982). The Numic Spread: Great
Basin Cultures in Competition. \emph{American Antiquity}, \emph{47}(3),
485--503. \url{https://doi.org/10.2307/280231}

\bibitem[\citeproctext]{ref-blessing2021}
Blessing, M. A., \& Schmidt, P. (2021). On the efficiency of
palaeolithic birch tar making. \emph{Journal of Archaeological Science:
Reports}, \emph{38}, 103096.
\url{https://doi.org/10.1016/j.jasrep.2021.103096}

\bibitem[\citeproctext]{ref-broman2018}
Broman, K. W., \& Woo, K. H. (2018). Data organization in spreadsheets.
\emph{The American Statistician}, \emph{72}(1), 2--10.
\url{https://doi.org/10.1080/00031305.2017.1375989}

\bibitem[\citeproctext]{ref-carroll2020}
Carroll, S. R., Garba, I., Figueroa-Rodríguez, O. L., Holbrook, J.,
Lovett, R., Materechera, S., Parsons, M., Raseroka, K.,
Rodriguez-Lonebear, D., Rowe, R., Sara, R., Walker, J. D., Anderson, J.,
\& Hudson, M. (2020). The CARE Principles for Indigenous Data
Governance. \emph{Data Science Journal}, \emph{19}(1), 43.
\url{https://doi.org/10.5334/dsj-2020-043}

\bibitem[\citeproctext]{ref-clay2018}
Clay, Z., \& Tennie, C. (2018). Is Overimitation a Uniquely Human
Phenomenon? Insights From Human Children as Compared to Bonobos.
\emph{Child Development}, \emph{89}(5), 1535--1544.
\url{https://doi.org/10.1111/cdev.12857}

\bibitem[\citeproctext]{ref-coles1979}
Coles, J. M. (1979). \emph{Experimental archaeology}. Academic Press.

\bibitem[\citeproctext]{ref-conrad2023}
Conrad, G., Hough, S., Baldino, J., Gala, N., Buchanan, B., Walker, R.
S., Key, A., Redmond, B. G., Bebber, M. R., \& Eren, M. I. (2023).
Clovis bone versus stone weapon tip penetration: Thinking about relative
costs and benefits, experimental assumptions, and archaeological
unknowns at sheriden cave, ohio, u.s.a. \emph{Journal of Archaeological
Science: Reports}, \emph{52}, 104295.
\url{https://doi.org/10.1016/j.jasrep.2023.104295}

\bibitem[\citeproctext]{ref-cordoni2022}
Cordoni, G., Pirarba, L., Elies, S., Demuru, E., Guéry, J.-P., \&
Norscia, I. (2022). Adult{\textendash}adult play in captive lowland
gorillas (Gorilla gorilla gorilla). \emph{Primates}, \emph{63}(3),
225--235. \url{https://doi.org/10.1007/s10329-022-00973-7}

\bibitem[\citeproctext]{ref-crabtree1977}
Crabtree, D. E. (1977). \emph{The obtuse angle as a functional edge} (D.
Ingersoll, J. E. Yellen, \& W. MacDonald, Eds.; pp. 38--51). Columbia
University Press.

\bibitem[\citeproctext]{ref-creswell2017}
Creswell, J. W., \& Clark, V. L. P. (2017). \emph{Designing and
conducting mixed methods research} (3rd edition). SAGE Publications,
Inc.

\bibitem[\citeproctext]{ref-cristino2010}
Cristino, F., Mathôt, S., Theeuwes, J., \& Gilchrist, I. D. (2010).
ScanMatch: A novel method for comparing fixation sequences.
\emph{Behavior Research Methods}, \emph{42}(3), 692--700.
\url{https://doi.org/10.3758/BRM.42.3.692}

\bibitem[\citeproctext]{ref-cueva-temprana2019}
Cueva-Temprana, A., Lombao, D., Morales, J. I., Geribàs, N., \&
Mosquera, M. (2019). Gestures during knapping: A two-perspective
approach to pleistocene technologies. \emph{Lithic Technology},
\emph{44}(2), 74--89.
\url{https://doi.org/10.1080/01977261.2019.1587255}

\bibitem[\citeproctext]{ref-cunningham2021}
Cunningham, S. (2021). \emph{Causal inference: The mixtape}. Yale
University Press. \url{https://doi.org/10.2307/j.ctv1c29t27}

\bibitem[\citeproctext]{ref-currie2022}
Currie, A. (2022). Speculation Made Material: Experimental Archaeology
and Maker{'}s Knowledge. \emph{Philosophy of Science}, \emph{89}(2),
337--359. \url{https://doi.org/10.1017/psa.2021.31}

\bibitem[\citeproctext]{ref-david2004}
David, B. (2004). Intentionality, Agency and an Archaeology of Choice.
\emph{Cambridge Archaeological Journal}, \emph{14}(1), 67--71.
\url{https://doi.org/10.1017/S0959774304220054}

\bibitem[\citeproctext]{ref-day2013}
Day, J. (2013). \emph{Making senses of the past: Toward a sensory
archaeology}. Southern Illinois University Press.
\url{https://muse.jhu.edu/pub/186/monograph/book/23997}

\bibitem[\citeproctext]{ref-delage2017}
Delage, C. (2017). Once upon a time{\ldots{}}the (hi)story of the
concept of the chaîne opératoire in french prehistory. \emph{World
Archaeology}, \emph{49}(2), 158173.
\url{https://doi.org/10.1080/00438243.2017.1300104}

\bibitem[\citeproctext]{ref-dobres1999}
Dobres, M.-A. (1999). \emph{Technology{'}s links and chaînes: The
processual unfolding of technique and technician} (D. Marcia-Anne \& C.
R. Hoffman, Eds.; p. 124146). Smithsonian Institution Press.
\url{http://ww.plndp.org/Departments/Joukowsky_Institute/courses/archaeologicaltheory/files/3787748.pdf}

\bibitem[\citeproctext]{ref-domuxednguez-rodrigo2008}
Domínguez-Rodrigo, M. (2008). Conceptual premises in experimental design
and their bearing on the use of analogy: An example from experiments on
cut marks. \emph{World Archaeology}, \emph{40}(1), 67--82.
\url{https://doi.org/10.1080/00438240701843629}

\bibitem[\citeproctext]{ref-douglass2020}
Douglass, K. (2020). Amy ty lilin-draza{'}ay: Building Archaeological
Practice on Principles of Community. \emph{African Archaeological
Review}, \emph{37}(3), 481--485.
\url{https://doi.org/10.1007/s10437-020-09404-8}

\bibitem[\citeproctext]{ref-erenExperimentalBisonButchery2024}
Eren, M. I., Bebber, M. R., Mukusha, L., Wilson, M., Boehm, A. R.,
Buchanan, B., Miller, G. L., Skoglund, M., Hayes, J., Barta, M., Bates,
S., Callaghan, R., Floyd, C., Morris, S., Neuharth, S., Newcomb, C.,
Rinella, S., Schneider, C., Smith, M. M., \ldots{} Meltzer, D. J.
(2024). Experimental bison butchery using replica hafted {Clovis} fluted
points and large handheld flakes. \emph{Journal of Archaeological
Science: Reports}, 104480.
\url{https://doi.org/10.1016/j.jasrep.2024.104480}

\bibitem[\citeproctext]{ref-eren2016}
Eren, M. I., Lycett, S. J., Patten, R. J., Buchanan, B., Pargeter, J.,
\& O'Brien, M. J. (2016). Test, model, and method validation: The role
of experimental stone artifact replication in hypothesis-driven
archaeology. \emph{Ethnoarchaeology: Journal of Archaeological,
Ethnographic and Experimental Studies}, \emph{8}(2), 103--136.
\url{https://doi.org/10.1080/19442890.2016.1213972}

\bibitem[\citeproctext]{ref-eren2024}
Eren, M. I., \& Meltzer, D. J. (2024). Controls, conceits, and aiming
for robust inferences in experimental archaeology. \emph{Journal of
Archaeological Science: Reports}, \emph{53}, 104411.
\url{https://doi.org/10.1016/j.jasrep.2024.104411}

\bibitem[\citeproctext]{ref-faniel2018}
Faniel, I. M., Austin, A., Kansa, E., Kansa, S. W., France, P., Jacobs,
J., Boytner, R., \& Yakel, E. (2018). Beyond the Archive: Bridging Data
Creation and Reuse in Archaeology. \emph{Advances in Archaeological
Practice}, \emph{6}(2), 105--116.
\url{https://doi.org/10.1017/aap.2018.2}

\bibitem[\citeproctext]{ref-flenniken1984}
Flenniken, J. J. (1984). The past, present, and future of flintknapping:
An anthropological perspective. \emph{Annual Review of Anthropology},
\emph{13}(1), 187--203.
\url{https://doi.org/10.1146/annurev.an.13.100184.001155}

\bibitem[\citeproctext]{ref-fragaszy2018}
Fragaszy, D. M., \& Mangalam, M. (2018). \emph{Chapter Five - Tooling}
(M. Naguib, L. Barrett, S. D. Healy, J. Podos, L. W. Simmons, \& M. Zuk,
Eds.; Vol. 50, pp. 177--241). Academic Press.
\url{https://doi.org/10.1016/bs.asb.2018.01.001}

\bibitem[\citeproctext]{ref-friard2016}
Friard, O., \& Gamba, M. (2016). BORIS: a free, versatile open-source
event-logging software for video/audio coding and live observations.
\emph{Methods in Ecology and Evolution}, \emph{7}(11), 1325--1330.
\url{https://doi.org/10.1111/2041-210X.12584}

\bibitem[\citeproctext]{ref-gergely2006}
Gergely, G., \& Csibra, G. (2006). \emph{Sylvia's recipe: The role of
imitation and pedagogy in the transmission of cultural knowledge} (S. C.
Levinson \& N. J. Enfield, Eds.; pp. 229--255). Berg Publishers.

\bibitem[\citeproctext]{ref-ghirlanda2006}
Ghirlanda, S., Enquist, M., \& Nakamaru, M. (2006). Cultural evolution
develops its own rules: The rise of conservatism and persuasion.
\emph{Current Anthropology}, \emph{47}(6), 1027--1034.
\url{https://doi.org/10.1086/508696}

\bibitem[\citeproctext]{ref-gilmore2015}
Gilmore, R., Adolph, K., Millman, D., Steiger, L., \& Simon, D. (2015).
Sharing displays and data from vision science research with databrary.
\emph{Journal of Vision}, \emph{15}(12), 280.
\url{https://doi.org/10.1167/15.12.280}

\bibitem[\citeproctext]{ref-gowlland2019}
Gowlland, G. (2019). The sociality of enskilment. \emph{Ethnos},
\emph{84}(3), 508--524.
\url{https://doi.org/10.1080/00141844.2018.1455726}

\bibitem[\citeproctext]{ref-griffin2013}
Griffin, D., Freedman, D. L., Nicholson, B., McConachie, F., \&
Parmington, A. (2013). The koorong project: Experimental archaeology and
wurundjeri continuation of cultural practices. \emph{Excavations,
Surveys and Heritage Management in Victoria}, \emph{2}, 5965.

\bibitem[\citeproctext]{ref-haidleHowThinkSimple2009}
Haidle, M. N. (2009). How to think a simple spear. In S. A. de Beaune,
F. L. Coolidge, \& T. Wynn (Eds.), \emph{Cognitive archaeology and human
evolution} (pp. 57--73). Cambridge University Press.

\bibitem[\citeproctext]{ref-haidle2010}
Haidle, M. N. (2010). Working-memory capacity and the evolution of
modern cognitive potential: Implications from animal and early human
tool use. \emph{Current Anthropology}, \emph{51}(S1), S149--S166.
\url{https://doi.org/10.1086/650295}

\bibitem[\citeproctext]{ref-haidle2014}
Haidle, M. N. (2014). Building a bridge{\textemdash}an archeologist's
perspective on the evolution of causal cognition. \emph{Frontiers in
Psychology}, \emph{5}.
\url{https://www.frontiersin.org/articles/10.3389/fpsyg.2014.01472}

\bibitem[\citeproctext]{ref-haidle2023}
Haidle, M. N. (2023). \emph{Cognigrams: Systematically reconstructing
behavioral architectures as a basis for cognitive archaeology} (T. Wynn,
K. A. Overmann, \& F. L. Coolidge, Eds.; p. C12S1C12S8). Oxford
University Press.
\url{https://doi.org/10.1093/oxfordhb/9780192895950.013.12}

\bibitem[\citeproctext]{ref-harris2021}
Harris, J. A., Boyd, R., \& Wood, B. M. (2021). The role of causal
knowledge in the evolution of traditional technology. \emph{Current
Biology}, \emph{31}(8), 1798--1803.e3.
\url{https://doi.org/10.1016/j.cub.2021.01.096}

\bibitem[\citeproctext]{ref-haythorn2018}
Haythorn, R., Buchanan, B., \& Eren, M. I. (2018). A new look at flaked
stone projectiles from the mixter site (33-ER-4), erie county, ohio,
USA. \emph{Lithic Technology}, \emph{43}(3), 166171.
\url{https://doi.org/10.1080/01977261.2018.1479950}

\bibitem[\citeproctext]{ref-henrichSearchHomoEconomicus2001}
Henrich, J., Boyd, R., Bowles, S., Camerer, C., Fehr, E., Gintis, H., \&
McElreath, R. (2001). In {Search} of {Homo} {Economicus}: {Behavioral}
{Experiments} in 15 {Small}-{Scale} {Societies}. \emph{American Economic
Review}, \emph{91}(2), 73--78. \url{https://doi.org/10.1257/aer.91.2.73}

\bibitem[\citeproctext]{ref-hernan2023}
Hernan, M. A., \& Robins, J. M. (2023). \emph{Causal inference: What
if}. CRC Press.

\bibitem[\citeproctext]{ref-heyes2012}
Heyes, C. (2012). Simple minds: A qualified defence of associative
learning. \emph{Philosophical Transactions of the Royal Society B:
Biological Sciences}, \emph{367}(1603), 2695--2703.
\url{https://doi.org/10.1098/rstb.2012.0217}

\bibitem[\citeproctext]{ref-hinds1999}
Hinds, P. J. (1999). The curse of expertise: The effects of expertise
and debiasing methods on prediction of novice performance. \emph{Journal
of Experimental Psychology: Applied}, \emph{5}, 205--221.
\url{https://doi.org/10.1037/1076-898X.5.2.205}

\bibitem[\citeproctext]{ref-hiscock2004}
Hiscock, P. (2004). Slippery and Billy: Intention, Selection and
Equifinality in Lithic Artefacts. \emph{Cambridge Archaeological
Journal}, \emph{14}(1), 71--77.
\url{https://doi.org/10.1017/S0959774304230050}

\bibitem[\citeproctext]{ref-holleman2020}
Holleman, G. A., Hooge, I. T., Kemner, C., \& Hessels, R. S. (2020). The
{`}real-world approach{'}and its problems: A critique of the term
ecological validity. \emph{Frontiers in Psychology}, \emph{11}, 721.

\bibitem[\citeproctext]{ref-horner2005}
Horner, V., \& Whiten, A. (2005). Causal knowledge and
imitation/emulation switching in chimpanzees (Pan troglodytes) and
children (Homo sapiens). \emph{Animal Cognition}, \emph{8}(3), 164--181.
\url{https://doi.org/10.1007/s10071-004-0239-6}

\bibitem[\citeproctext]{ref-huggett2018}
Huggett, J. (2018). Reuse Remix Recycle: Repurposing Archaeological
Digital Data. \emph{Advances in Archaeological Practice}, \emph{6}(2),
93--104. \url{https://doi.org/10.1017/aap.2018.1}

\bibitem[\citeproctext]{ref-hussain2021}
Hussain, S. T., \& Will, M. (2021). Materiality, Agency and Evolution of
Lithic Technology: an Integrated Perspective for Palaeolithic
Archaeology. \emph{Journal of Archaeological Method and Theory},
\emph{28}(2), 617--670. \url{https://doi.org/10.1007/s10816-020-09483-6}

\bibitem[\citeproctext]{ref-johnson1978}
Johnson, L. L. (1978). A history of flint-knapping experimentation,
1838-1976 {[}and comments and reply{]}. \emph{Current Anthropology},
\emph{19}(2), 337--372. \url{https://doi.org/10.1086/202078}

\bibitem[\citeproctext]{ref-kafaee2022}
Kafaee, M., Daviran, E., \& Taqavi, M. (2022). The QWERTY keyboard from
the perspective of the Collingridge dilemma: lessons for co-construction
of human-technology. \emph{AI \& SOCIETY}.
\url{https://doi.org/10.1007/s00146-022-01573-1}

\bibitem[\citeproctext]{ref-lasalle2010}
La Salle, M. J. (2010). Community Collaboration and Other Good
Intentions. \emph{Archaeologies}, \emph{6}(3), 401--422.
\url{https://doi.org/10.1007/s11759-010-9150-8}

\bibitem[\citeproctext]{ref-li2022}
Li, L., Lin, S. C., McPherron, S. P., Abdolahzadeh, A., Chan, A.,
Dogandžić, T., Iovita, R., Leader, G. M., Magnani, M., Rezek, Z., \&
Dibble, H. L. (2022). A Synthesis of the Dibble et al. Controlled
Experiments into the Mechanics of Lithic Production. \emph{Journal of
Archaeological Method and Theory}.
\url{https://doi.org/10.1007/s10816-022-09586-2}

\bibitem[\citeproctext]{ref-lin2018}
Lin, S. C., Rezek, Z., \& Dibble, H. L. (2018). Experimental Design and
Experimental Inference in Stone Artifact Archaeology. \emph{Journal of
Archaeological Method and Theory}, \emph{25}(3), 663--688.
\url{https://doi.org/10.1007/s10816-017-9351-1}

\bibitem[\citeproctext]{ref-liu2023}
Liu, C., Khreisheh, N., Stout, D., \& Pargeter, J. (2023). Differential
effects of knapping skill acquisition on the cultural reproduction of
Late Acheulean handaxe morphology: Archaeological and experimental
insights. \emph{Journal of Archaeological Science: Reports}, \emph{49},
103974. \url{https://doi.org/10.1016/j.jasrep.2023.103974}

\bibitem[\citeproctext]{ref-liuInferringCulturalReproduction2023}
Liu, C., \& Stout, D. (2023). Inferring cultural reproduction from
lithic data: {A} critical review. \emph{Evolutionary Anthropology:
Issues, News, and Reviews}, \emph{32}(2), 83--99.
\url{https://doi.org/10.1002/evan.21964}

\bibitem[\citeproctext]{ref-lombao2017}
Lombao, D., Guardiola, M., \& Mosquera, M. (2017). Teaching to make
stone tools: new experimental evidence supporting a technological
hypothesis for the origins of language. \emph{Scientific Reports},
\emph{7}(1), 14394. \url{https://doi.org/10.1038/s41598-017-14322-y}

\bibitem[\citeproctext]{ref-lyons2007}
Lyons, D. E., Young, A. G., \& Keil, F. C. (2007). The hidden structure
of overimitation. \emph{Proceedings of the National Academy of Sciences
of the United States of America}, \emph{104}(50), 19751--19756.
\url{https://doi.org/10.1073/pnas.0704452104}

\bibitem[\citeproctext]{ref-mahaney2014}
Mahaney, R. A. (2014). Exploring the complexity and structure of
acheulean stoneknapping in relation to natural language.
\emph{PaleoAnthropology}, \emph{2014}, 586606.
\url{https://doi.org/10.4207/PA.2014.ART90}

\bibitem[\citeproctext]{ref-malafouris2013}
Malafouris, L. (2013). \emph{How things shape the mind: A theory of
material engagement}. The MIT Press.

\bibitem[\citeproctext]{ref-maloney2020}
Maloney, T. R., \& Street, M. (2020). Hot debate: Identifying heat
treatment in Australian archaeology using science and modern indigenous
knowledge. \emph{Quaternary Science Reviews}, \emph{241}, 106431.
\url{https://doi.org/10.1016/j.quascirev.2020.106431}

\bibitem[\citeproctext]{ref-marreiros2020}
Marreiros, J., Pereira, T., \& Iovita, R. (2020). Controlled experiments
in lithic technology and function. \emph{Archaeological and
Anthropological Sciences}, \emph{12}(6), 110.
\url{https://doi.org/10.1007/s12520-020-01059-5}

\bibitem[\citeproctext]{ref-marshall2002}
Marshall, Y. (2002). What is community archaeology? \emph{World
Archaeology}, \emph{34}(2), 211--219.
\url{https://doi.org/10.1080/0043824022000007062}

\bibitem[\citeproctext]{ref-martellotta2022}
Martellotta, E. F., Perston, Y. L., Craft, P., Wilkins, J., \& Langley,
M. C. (2022). Beyond the main function: An experimental study of the use
of hardwood boomerangs in retouching activities. \emph{PLOS ONE},
\emph{17}(8), e0273118.
\url{https://doi.org/10.1371/journal.pone.0273118}

\bibitem[\citeproctext]{ref-marwick2018}
Marwick, B., \& Birch, S. E. P. (2018). A Standard for the Scholarly
Citation of Archaeological Data as an Incentive to Data Sharing.
\emph{Advances in Archaeological Practice}, \emph{6}(2), 125--143.
\url{https://doi.org/10.1017/aap.2018.3}

\bibitem[\citeproctext]{ref-mesoudi2008}
Mesoudi, A., \& O'Brien, M. J. (2008). The cultural transmission of
great basin projectile-point technology i: An experimental simulation.
\emph{American Antiquity}, \emph{73}(1), 328.
https://doi.org/\url{https://doi.org/10.1017/S0002731600041263}

\bibitem[\citeproctext]{ref-milks2023}
Milks, A., Hoggard, C., \& Pope, M. (2023). Reassessing the
Interpretative Potential of Ethnographic Collections for Early Hunting
Technologies. \emph{Journal of Archaeological Method and Theory}.
\url{https://doi.org/10.1007/s10816-023-09635-4}

\bibitem[\citeproctext]{ref-armstrong2018}
Mindermann, S., \& Armstrong, S. (2018). Occam's razor is insufficient
to infer the preferences of irrational agents. \emph{Proceedings of the
32nd {International} {Conference} on {Neural} {Information} {Processing}
{Systems}}, 5603--5614.

\bibitem[\citeproctext]{ref-mobbs2021}
Mobbs, D., Wise, T., Suthana, N., Guzmán, N., Kriegeskorte, N., \&
Leibo, J. Z. (2021). Promises and challenges of human computational
ethology. \emph{Neuron}, \emph{109}(14), 2224--2238.
\url{https://doi.org/10.1016/j.neuron.2021.05.021}

\bibitem[\citeproctext]{ref-montgomery2023}
Montgomery, L. M., \& Fryer, T. C. (2023). The future of archaeology is
(still) community collaboration. \emph{Antiquity}, \emph{97}(394),
795--809. \url{https://doi.org/10.15184/aqy.2023.98}

\bibitem[\citeproctext]{ref-moody2021}
Moody, B., Dye, T., May, K., Wright, H., \& Buck, C. (2021). Digital
chronological data reuse in archaeology: Three case studies with varying
purposes and perspectives. \emph{Journal of Archaeological Science:
Reports}, \emph{40}, 103188.
\url{https://doi.org/10.1016/j.jasrep.2021.103188}

\bibitem[\citeproctext]{ref-moore2020}
Moore, M. W. (2020). Hominin Stone Flaking and the Emergence of
{`}Top-down{'} Design in Human Evolution. \emph{Cambridge Archaeological
Journal}, \emph{30}(4), 647--664.
\url{https://doi.org/10.1017/S0959774320000190}

\bibitem[\citeproctext]{ref-morin2022}
Morin, O. (2022). Cultural Conservatism. \emph{Journal of Cognition and
Culture}, \emph{22}(5), 406--420.
\url{https://doi.org/10.1163/15685373-12340142}

\bibitem[\citeproctext]{ref-muller2017}
Muller, A., Clarkson, C., \& Shipton, C. (2017). Measuring behavioural
and cognitive complexity in lithic technology throughout human
evolution. \emph{Journal of Anthropological Archaeology}, \emph{48},
166--180. \url{https://doi.org/10.1016/j.jaa.2017.07.006}

\bibitem[\citeproctext]{ref-muller2023}
Muller, A., Shipton, C., \& Clarkson, C. (2023). The Proceduralization
of Hominin Knapping Skill: Memorizing Different Lithic Technologies.
\emph{Cambridge Archaeological Journal}, 1--18.
\url{https://doi.org/10.1017/S0959774323000070}

\bibitem[\citeproctext]{ref-nami2010}
Nami, H., G. (2010). Theoretical {Reflections} on {Experimental}
{Archaeology} and {Lithic} {Technology}: {Issues} on {Actualistic}
{Stone} {Tools} {Analysis} and {Interpretation}. In H. Nami G. (Ed.),
\emph{Experiments and {Interpretation} of {Traditional} {Technologies}:
{Essays} in {Honor} of {Errett} {Callahan}} (pp. 91--168). Ediciones de
ArqueologÌa Contempornea.

\bibitem[\citeproctext]{ref-nastase2020}
Nastase, S. A., Goldstein, A., \& Hasson, U. (2020). Keep it real:
rethinking the primacy of experimental control in cognitive
neuroscience. \emph{NeuroImage}, \emph{222}, 117254.
\url{https://doi.org/10.1016/j.neuroimage.2020.117254}

\bibitem[\citeproctext]{ref-nielsen2014}
Nielsen, M., Mushin, I., Tomaselli, K., \& Whiten, A. (2014). Where
Culture Takes Hold: {``}Overimitation{''} and Its Flexible Deployment in
Western, Aboriginal, and Bushmen Children. \emph{Child Development},
\emph{85}(6), 2169--2184. \url{https://doi.org/10.1111/cdev.12265}

\bibitem[\citeproctext]{ref-nielsenFailureFindOverimitation2010}
Nielsen, M., \& Susianto, E. W. E. (2010). Failure to find
over-imitation in captive orangutans ({Pongo} pygmaeus): {Implications}
for our understanding of cross-generation information transfer. In J.
Håkansson (Ed.), \emph{Developmental {Psychology}} (pp. 153--167). Nova
Science Publishers.

\bibitem[\citeproctext]{ref-nielsen2010}
Nielsen, M., \& Tomaselli, K. (2010). Overimitation in Kalahari Bushman
Children and the Origins of Human Cultural Cognition.
\emph{Psychological Science}, \emph{21}(5), 729--736.
\url{https://doi.org/10.1177/0956797610368808}

\bibitem[\citeproctext]{ref-oneill2019}
O'Neill, B., \& O'Sullivan, A. (2019). \emph{Experimental archaeology
and (re)-experiencing the senses of the medieval world} (R. Skeates \&
J. Day, Eds.; pp. 451--466). Routledge.

\bibitem[\citeproctext]{ref-outram2008}
Outram, A. K. (2008). Introduction to experimental archaeology.
\emph{World Archaeology}, \emph{40}(1), 1--6.
\url{https://www.jstor.org/stable/40025310}

\bibitem[\citeproctext]{ref-ouzman2023}
Ouzman, S. (2023). Authorship, attribution and acknowledgment in
archaeology. \emph{Australian Archaeology}, \emph{89}(1), 6670.
\url{https://doi.org/10.1080/03122417.2023.2190497}

\bibitem[\citeproctext]{ref-pargeter2023}
Pargeter, J., Liu, C., Kilgore, M. B., Majoe, A., \& Stout, D. (2023).
Testing the Effect of Learning Conditions and Individual Motor/Cognitive
Differences on Knapping Skill Acquisition. \emph{Journal of
Archaeological Method and Theory}, \emph{30}(1), 127--171.
\url{https://doi.org/10.1007/s10816-022-09592-4}

\bibitem[\citeproctext]{ref-perreault2013}
Perreault, C., Brantingham, P. J., Kuhn, S. L., Wurz, S., \& Gao, X.
(2013). Measuring the complexity of lithic technology. \emph{Current
Anthropology}, \emph{54}(S8), S397--S406.
\url{https://doi.org/10.1086/673264}

\bibitem[\citeproctext]{ref-pfleging2019}
Pfleging, J., Iovita, R., \& Buchli, J. (2019). Influence of force and
duration on stone tool wear: results from experiments with a
force-controlled robot. \emph{Archaeological and Anthropological
Sciences}, \emph{11}(11), 5921--5935.
\url{https://doi.org/10.1007/s12520-018-0729-0}

\bibitem[\citeproctext]{ref-porqueddu2023}
Porqueddu, M.-E., Sciuto, C., \& Lamesa, A. (2023). Reconsidering the
Chaîne Opératoire: At the Crossroad Between People and Materials.
\emph{Open Archaeology}, \emph{9}(1).
\url{https://doi.org/10.1515/opar-2022-0296}

\bibitem[\citeproctext]{ref-premoEquifinalityExplanationThoughts2010}
Premo, L. S. (2010). Equifinality and explanation: {Thoughts} on the
role of agent-based modeling in postpositivist archaeology. In A.
Costopoulos \& M. W. Lake (Eds.), \emph{Simulating {Change}:
{Archaeology} {Into} the {Twenty}-first {Century}} (pp. 28--37).
University of Utah Press.

\bibitem[\citeproctext]{ref-proffitt2022}
Proffitt, T., Bargalló, A., \& Torre, I. de la. (2022). The Effect of
Raw Material on the Identification of Knapping Skill: a Case Study from
Olduvai Gorge, Tanzania. \emph{Journal of Archaeological Method and
Theory}, \emph{29}(1), 50--82.
\url{https://doi.org/10.1007/s10816-021-09511-z}

\bibitem[\citeproctext]{ref-quillien2021}
Quillien, T., \& German, T. C. (2021). A simple definition of
{`}intentionally{'}. \emph{Cognition}, \emph{214}, 104806.
\url{https://doi.org/10.1016/j.cognition.2021.104806}

\bibitem[\citeproctext]{ref-ranhorn2020}
Ranhorn, K. L., Pargeter, J., \& Premo, L. S. (2020). Investigating the
evolution of human social learning through collaborative experimental
archaeology. \emph{Evolutionary Anthropology: Issues, News, and
Reviews}, \emph{29}(2), 53--55. \url{https://doi.org/10.1002/evan.21823}

\bibitem[\citeproctext]{ref-reeves2009}
Reeves, D., Bury, R., \& Robinson, D. W. (2009). Invoking occam's razor:
Experimental pigment processing and an hypothesis concerning emigdiano
chumash rock art. \emph{Journal of California and Great Basin
Anthropology}, \emph{29}(1), 59--67.
\url{https://www.jstor.org/stable/27825902}

\bibitem[\citeproctext]{ref-reevesflores2012}
Reeves Flores, J. (2012). \emph{Experimental archaeology: an ethnography
of its perceived value and impact in archaeological research} {[}PhD
thesis{]}. \url{https://ore.exeter.ac.uk/repository/handle/10871/9041}

\bibitem[\citeproctext]{ref-reid1975}
Reid, J. J., Schiffer, M. B., \& Rathje, W. L. (1975). Behavioral
Archaeology: Four Strategies. \emph{American Anthropologist},
\emph{77}(4), 864--869.
\url{https://doi.org/10.1525/aa.1975.77.4.02a00090}

\bibitem[\citeproctext]{ref-reynolds1999}
Reynolds, P. J. (1999). The nature of experiment in archaeology. In A.
Harding (Ed.), \emph{Experiment and design: {Archaeological} studies in
honour of {John} {Coles}} (pp. 156--162). Oxbow Books.

\bibitem[\citeproctext]{ref-roe2009}
Roe, B. E., \& Just, D. R. (2009). Internal and external validity in
economics research: Tradeoffs between experiments, field experiments,
natural experiments, and field data. \emph{American Journal of
Agricultural Economics}, \emph{91}(5), 1266--1271.
\url{https://www.jstor.org/stable/20616293}

\bibitem[\citeproctext]{ref-roepstorff2008}
Roepstorff, A. (2008). Things to think with: Words and objects as
material symbols. \emph{Philosophical Transactions of the Royal Society
B: Biological Sciences}, \emph{363}(1499), 2049--2054.
\url{https://doi.org/10.1098/rstb.2008.0015}

\bibitem[\citeproctext]{ref-roux2005}
Roux, V., \& David, É. (2005). \emph{Planning abilities as a dynamic
perceptual-motor skill: an actualist study of different levels of
expertise involved in stone knapping} (V. Roux \& B. Bril, Eds.; pp.
91--108). McDonald Institute for Archaeological Research.
\url{https://shs.hal.science/halshs-00120262}

\bibitem[\citeproctext]{ref-russell2004}
Russell, L. (2004). Drinking from the Penholder: Intentionality and
Archaeological Theory. \emph{Cambridge Archaeological Journal},
\emph{14}(1), 64--67. \url{https://doi.org/10.1017/S0959774304210058}

\bibitem[\citeproctext]{ref-schiffer2010}
Schiffer, M. B. (2010). \emph{Behavioral Archaeology: Principles and
Practice}. Routledge.

\bibitem[\citeproctext]{ref-schillinger2016}
Schillinger, K., Mesoudi, A., \& Lycett, S. J. (2016). Copying error,
evolution, and phylogenetic signal in artifactual traditions: An
experimental approach using {``}model artifacts{''}. \emph{Journal of
Archaeological Science}, \emph{70}, 23--34.
\url{https://doi.org/10.1016/j.jas.2016.04.013}

\bibitem[\citeproctext]{ref-schmidt2019}
Schmidt, P., Blessing, M., Rageot, M., Iovita, R., Pfleging, J., Nickel,
K. G., Righetti, L., \& Tennie, C. (2019). Birch tar production does not
prove Neanderthal behavioral complexity. \emph{Proceedings of the
National Academy of Sciences}, \emph{116}(36), 17707--17711.
\url{https://doi.org/10.1073/pnas.1911137116}

\bibitem[\citeproctext]{ref-schmidt2020}
Schmidt, S. C., \& Marwick, B. (2020). Tool-Driven Revolutions in
Archaeological Science. \emph{Journal of Computer Applications in
Archaeology}, \emph{3}(1), 1832. \url{https://doi.org/10.5334/jcaa.29}

\bibitem[\citeproctext]{ref-schneider2020}
Schneider, T. D., \& Hayes, K. (2020). Epistemic colonialism: Is it
possible to decolonize archaeology? \emph{American Indian Quarterly},
\emph{44}(2), 127--148.
\url{https://doi.org/10.5250/amerindiquar.44.2.0127}

\bibitem[\citeproctext]{ref-shamay-tsoory2019}
Shamay-Tsoory, S. G., \& Mendelsohn, A. (2019). Real-Life Neuroscience:
An Ecological Approach to Brain and Behavior Research.
\emph{Perspectives on Psychological Science}, \emph{14}(5), 841--859.
\url{https://doi.org/10.1177/1745691619856350}

\bibitem[\citeproctext]{ref-simon2015}
Simon, D. A., Gordon, A. S., Steiger, L., \& Gilmore, R. O. (2015).
\emph{Databrary: Enabling sharing and reuse of research video}. 279280.
\url{https://doi.org/10.1145/2756406.2756951}

\bibitem[\citeproctext]{ref-therout2019}
Skeates, R., \& Day, J. (Eds.). (2019). \emph{The routledge handbook of
sensory archaeology}. Routledge.
\url{https://doi.org/10.4324/9781315560175}

\bibitem[\citeproctext]{ref-sonkusare2019}
Sonkusare, S., Breakspear, M., \& Guo, C. (2019). Naturalistic Stimuli
in Neuroscience: Critically Acclaimed. \emph{Trends in Cognitive
Sciences}, \emph{23}(8), 699--714.
\url{https://doi.org/10.1016/j.tics.2019.05.004}

\bibitem[\citeproctext]{ref-soressi2011}
Soressi, M., \& Geneste, J.-M. (2011). The History and Efficacy of the
Chaîne Opératoire Approach to Lithic Analysis: Studying Techniques to
Reveal Past Societies in an Evolutionary Perspective.
\emph{PaleoAnthropology}, \emph{2011}, 334--350.
\url{https://paleoanthropology.org/ojs/index.php/paleo/article/view/643}

\bibitem[\citeproctext]{ref-stengelin2020}
Stengelin, R., Hepach, R., \& Haun, D. B. M. (2020). Cross-cultural
variation in how much, but not whether, children overimitate.
\emph{Journal of Experimental Child Psychology}, \emph{193}, 104796.
\url{https://doi.org/10.1016/j.jecp.2019.104796}

\bibitem[\citeproctext]{ref-stout2002}
Stout, D. (2002). Skill and cognition in stone tool production: An
ethnographic case study from irian jaya. \emph{Current Anthropology},
\emph{43}(5), 693--722. \url{https://doi.org/10.1086/342638}

\bibitem[\citeproctext]{ref-stoutCognitiveScienceTechnology2021}
Stout, D. (2021). The cognitive science of technology. \emph{Trends in
Cognitive Sciences}, \emph{25}(11), 964--977.
\url{https://doi.org/10.1016/j.tics.2021.07.005}

\bibitem[\citeproctext]{ref-stout2021}
Stout, D., Chaminade, T., Apel, J., Shafti, A., \& Faisal, A. A. (2021).
The measurement, evolution, and neural representation of action grammars
of human behavior. \emph{Scientific Reports}, \emph{11}(1).
\url{https://doi.org/10.1038/s41598-021-92992-5}

\bibitem[\citeproctext]{ref-subiaul2016}
Subiaul, F., Winters, K., Krumpak, K., \& Core, C. (2016). Vocal
overimitation in preschool-age children. \emph{Journal of Experimental
Child Psychology}, \emph{141}, 145--160.
\url{https://doi.org/10.1016/j.jecp.2015.08.010}

\bibitem[\citeproctext]{ref-syed2022}
Syed, M., \& McLean, K. C. (2022). Disentangling paradigm and method can
help bring qualitative research to post-positivist psychology and
address the generalizability crisis. \emph{Behavioral and Brain
Sciences}, \emph{45}, e32.
\url{https://doi.org/10.1017/S0140525X21000431}

\bibitem[\citeproctext]{ref-timbrell2023}
Timbrell, L. (2023). A {Collaborative} {Model} for {Lithic} {Shape}
{Digitization} in {Museum} {Settings}. \emph{Lithic Technology},
\emph{48}(1), 31--42.
\url{https://doi.org/10.1080/01977261.2022.2092299}

\bibitem[\citeproctext]{ref-varela2017}
Varela, F. J., Thompson, E., \& Rosch, E. (2017). \emph{The {Embodied}
{Mind}: {Cognitive} {Science} and {Human} {Experience}} (revised
edition). The MIT Press.

\bibitem[\citeproctext]{ref-weger2019}
Weger, U. W., Wagemann, J., \& Tewes, C. (2019). Editorial: The
challenges and opportunities of introspection in psychology: Theory and
method. \emph{Frontiers in Psychology}, \emph{10}.
\url{https://www.frontiersin.org/articles/10.3389/fpsyg.2019.02196}

\bibitem[\citeproctext]{ref-whittaker1994}
Whittaker, J. C. (1994). \emph{Flintknapping: Making and Understanding
Stone Tools}. University of Texas Press.

\bibitem[\citeproctext]{ref-whittaker2004}
Whittaker, J. C. (2004). \emph{American Flintknappers: Stone Age Art in
the Age of Computers}. University of Texas Press.

\bibitem[\citeproctext]{ref-wilkinson2016}
Wilkinson, M. D., Dumontier, M., Aalbersberg, Ij. J., Appleton, G.,
Axton, M., Baak, A., Blomberg, N., Boiten, J.-W., Silva Santos, L. B.
da, Bourne, P. E., Bouwman, J., Brookes, A. J., Clark, T., Crosas, M.,
Dillo, I., Dumon, O., Edmunds, S., Evelo, C. T., Finkers, R., \ldots{}
Mons, B. (2016). The FAIR Guiding Principles for scientific data
management and stewardship. \emph{Scientific Data}, \emph{3}(1), 160018.
\url{https://doi.org/10.1038/sdata.2016.18}

\bibitem[\citeproctext]{ref-yarkoni2022}
Yarkoni, T. (2022). The generalizability crisis. \emph{Behavioral and
Brain Sciences}, \emph{45}, e1.
\url{https://doi.org/10.1017/S0140525X20001685}

\end{CSLReferences}

\end{document}

### 2024-05-13 manuscript/manuscript_word.Rmd
1st round revision submitted to Advances in Archaeological Practice.

@@
 ---
-title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
+title: "Variation matters: Expanding the scope of experimental archaeology"
 author: 
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
-abstract: "Abstract: This paper presents the Perception-Process-Product (\"Triple P\") framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
+abstract: "Abstract: This paper aims to expand the scope of experimental archaeology to emphasize multi-level variation and interactions across the levels of perception, actions, and outcomes.
+Such an approach, loosely formulated as the Perception-Process-Product (\"Triple P\") framework, offers a more grounded and richer explanation of the past archaeological record.
 It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+Serving as a heuristic device, this alternative mode of knowledge production is highly flexible in nature, where each single component is detachable as dictated by individual research questions.
 \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"
@@
 Over-imitation means the copying of actions that are causally irrelevant in a goal-directed action sequence [@lyons2007].
-It is a psychological propensity that was suggested to be uniquely prevalent among humans in inter-species [@horner2005; @clay2018] comparisons and cross-cultural contexts [@nielsen2010; @stengelin2020].
+It is a psychological propensity that was suggested to be uniquely prevalent among humans when compared with non-human primates including chimpanzees [@horner2005], bonobos [@clay2018], and orangutans [@nielsenFailureFindOverimitation2010].
+Subsequent research further suggested that within human societies over-imitation has been commonly observed among children across various cultural contexts [@nielsen2010; @stengelin2020; @nielsen2014; @subiaul2016; @berl2015].
 Gergely and Csisbra [-@gergely2006] introduced "Sylvia's Recipe" that vividly illustrates this cognitive process in the transmission of technical skills.
 Sylvia is an education researcher who developed a unique way of cooking ham roast by observing her mother during childhood, where she cut both ends of a ham.
-Later in life, her mother happened to watch her cooking, where she noticed and questioned the purpose of this step of preparation.
+Later in life, her mother happened to watch her cooking, during which she noticed and questioned the purpose of this step of preparation.
 Sylvia could not answer it and was then told that it was processed that way because her mother did not have a pan that was large enough to cook a full-sized ham.
@@
 
-Built upon this critique of *Homo economicus* as well as intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+Built upon this critique of *Homo economicus* as well as the four strategies^[Strategy 2 is particularly relevant here, which is "the pursuit of general questions in present-day material culture in order to acquiring laws for making useful behavioral inferences" (Schiffer, 2010: 6)] of behavioral archaeology [@schiffer2010; @reid1975], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
@@
 It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] ethnography.
-Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
+Likewise, novices' lack of flintknapping expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers.
 Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.
+Several researchers have also pointed out that literature-informed archaeologists sometimes get lost in reconstructing previous archaeologists’ reconstructions instead of searching for diverse solutions to better understand the actual archaeological phenomenon [@bell2014; @currie2022], which is another reason why we need the presence of hobbyists and novices in the community of experimental archaeology.
 
 Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artifacts.
-Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage.
+Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage [@arthur2024].
 This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022; @milks2023] across different regions.
@@
 This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research.
-To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.
+To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations [@ouzman2023].
+Moreover, considering the checkered disciplinary history of archaeology/anthropology featuring colonial exploitation, the changes in the evaluation system alone are not enough.
+This further highlights the need for adopting a community-based approach to fundamentally transform the power dynamics in archaeological knowledge production and distribution [@schneider2020; @lasalle2010; @atalay2012].
 
@@
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
-Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of technological complexity [@stout2021], or behavioral similarity across individuals [@mobbs2021; @cristino2010], etc.
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of the complexity of various technological systems, a classical topic in paleolithic archaeology and the evolution of human cognition [@perreault2013; @muller2017].
+Unlike the traditional approaches resorting to the extraction and comparison of lithic attributes, Stout et al.
+[-@stout2021] recorded the videos of expert flintknappers reproducing Oldowan and Acheulean technologies and then manually parsed their knapping activities using action grammar, generating multiple sequences of actions.
+Borrowing tools from computational linguistics, they then calculated the transition probability between each action category across two technological systems, which provided an objective and quantifiable proxy
+for measuring technological complexity.
+Another scenario of its application is the measurement of behavioral similarity across individuals [@mobbs2021; @cristino2010], which is particularly relevant in the above-mentioned cultural transmission experiments.
+Again, since the existing works on this topic mainly focus on the sole analysis of experimental replicas, many aspects of knapping skill learning processes remain unclear.
+For example, how do different individual learning strategies (high-fidelity action copying vs.
+predominantly trial-and-error learning) affect the morphological variation of their final products?
+Or will learning behavioral conformity within a community of practice necessarily lead to homogeneity in the formation of lithic assemblages?
+To answer these questions, the quantitative analysis of process-level data associated with the product-level data becomes necessary.
+Behatrix (**Figure** \@ref(fig:behatrix), <https://www.boris.unito.it/behatrix/>), a sister software of BORIS, allows us to calculate the action sequence (dis)similarity between novice learners and expert demonstrators/fellow novice learners using established algorithms [see @cordoni2022 for an application of analyzing play behavior similarity among gorillas].
 
@@
 
+```{r behatrix, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="The user interface of Behatrix displaying the transition probability between each action category in a handaxe knapping session."}
+knitr::include_graphics(here::here("figure/Fig3.jpg"))
+```
+
 ## Perception-level data
@@
 
-Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
+Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences [@day2013; @oneill2019; @therout2019], social interactions, and cultural meanings associated with the experimental activities [@gowlland2019].
 Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction.
-One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+One potential application of ethnographic methods in the experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
+For example, serial formation of step-fracture on the debitage surface is commonly interpreted as unintentional mistakes indicative of novice knappers, while in some cases researchers treat it as evidence of deliberate core rejuvenation [@akerman1993: 126].
+The accumulation of testimonies by participating knappers in terms of their intended outcome becomes useful in this scenario, although these materials should be examined in combination with the relevant product- and process-level data in a careful manner.
 Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely "an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen)." In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography becomes necessary.
 
-## Multi-level data curation
+## Multi-level sample and data curation
 
@@
 
+On top of the digital data curation, an easily ignored but crucial aspect regarding the integrity and reliability of research in experimental archaeology is the long-term and proper curation of psychical specimens produced during experiments, which is particularly relevant to the product-level data.
+Haythron et al.
+[-@haythorn2018] sharply pointed out that re-running statistical analyses on a publicly available spreadsheet containing incorrect lithic projectile attribute data would be meaningless.
+In this case, a reexamination of the actual experiment samples becomes necessary.
+Moreover, it is likely that new research questions can only be answered through direct observation and measurement of the experimental assemblages themselves instead of data readily available in previously compiled spreadsheets [@eren2016].
+The existence of these possibilities thus requires that an experimental assemblage of interest should be ideally curated in public institutions with easy access and detailed contextual information [@haythorn2018].
+
 # Conclusion
@@
 I thank Dietrich Stout for his helpful comments on earlier drafts of this article as well as Mark Moore and Shaya Jannati for inspiring discussions.
+I am also grateful to Metin Eren and two other anonymous reviewers for their insightful feedback.
 This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."

### 2024-07-25 manuscript/manuscript.Rmd
title and readme file update.

@@
 ---
-title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
+title: "Variation matters: Expanding the scope of experimental archaeology"
 author: 
@@
   ```{=latex}
-  This paper presents the Perception-Process-Product ("Triple P") framework that aims to expand the scope of experimental archaeology.
-The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record.
+  \textbf{Abstract:} This paper aims to expand the scope of experimental archaeology to emphasize multi-level variation and interactions across the levels of perception, actions, and outcomes.
+Such an approach, loosely formulated as the Perception-Process-Product ("Triple P") framework, offers a more grounded and richer explanation of the past archaeological record.
 It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.
+Serving as a heuristic device, this alternative mode of knowledge production is highly flexible in nature, where each single component is detachable as dictated by individual research questions.
 
   \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production
+
+  \textbf{Abstracto:} El objetivo de este artículo es ampliar el alcance de la arqueología experimental para centrar la atención en la variación multinivel y las interacciones entre los rangos de percepción, acciones y resultados.
+Este enfoque, formulado vagamente como el marco Percepción-Proceso-Producto ("Triple P"), ofrece una explicación más fundamentada y rica del registro arqueológico del pasado.
+Consta de tres principios 1) reconocer la compensación inherente entre control y generalizabilidad en el diseño de la investigación experimental; 2) fomentar proyectos de colaboración que impliquen a participantes en la investigación de procedencias distintas no limitados a los tradicionales participantes, como aficionados y novatos; 3) adoptar un flujo de trabajo que normalice la recogida y conservación de datos etológicos y etnográficos en proyectos experimentales.
+Como dispositivo heurístico, este modo alternativo de producción de conocimientos es muy flexible por naturaleza, y cada uno de sus componentes es desechable o modificable en función de las preguntas de investigación individuales.
+
+  \textbf{Palabras clave:} Arqueología experimental; Análisis etológico; Investigación etnográfica; Maldición del conocimiento; Producción colaborativa de conocimientos
   ```
@@
    - \usepackage{lineno}
+   - \usepackage[utf8]{inputenc}
    - \linenumbers
@@
 csl: apa.csl
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
+
+```{r, include=FALSE}
+options(tinytex.verbose = TRUE)
+```
 
@@
 
-Built upon this critique of *Homo economicus* as well as intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+Built upon this critique of *Homo economicus* as well as the four strategies$^1$ of behavioral archaeology [@schiffer2010; @reid1975], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
@@
 
-The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
+The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@degtiar2023; @roe2009: 1266-1267]$^2$.
 Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
-In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024].
+In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024; @marreiros2020].
 In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.).
@@
 
-Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @coles1979; @johnson1978; @whittaker1994: 54-61].
+Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the nineteenth century [@johnson1978; @whittaker1994: 54-61].
 Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers.
@@
 The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1).
-Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of technological complexity [@stout2021], or behavioral similarity across individuals [@mobbs2021; @cristino2010], etc.
+Depending on the original research question, sequences of coded actions can then be used in further analysis, such as the measurement of the complexity of various technological systems, a classical topic in paleolithic archaeology and the evolution of human cognition [@perreault2013; @muller2017].
+Unlike the traditional approaches resorting to the extraction and comparison of lithic attributes, Stout et al.
+[-@stout2021] recorded the videos of expert flintknappers reproducing Oldowan and Acheulean technologies and then manually parsed their knapping activities using action grammar, generating multiple sequences of actions.
+Borrowing tools from computational linguistics, they then calculated the transition probability between each action category across two technological systems, which provided an objective and quantifiable proxy for measuring technological complexity.
+Another scenario of its application is the measurement of behavioral similarity across individuals [@mobbs2021; @cristino2010], which is particularly relevant in the above-mentioned cultural transmission experiments.
+Again, since the existing works on this topic mainly focus on the sole analysis of experimental replicas, many aspects of knapping skill learning processes remain unclear.
+For example, how do different individual learning strategies (high-fidelity action copying vs.
+predominantly trial-and-error learning) affect the morphological variation of their final products?
+Or will learning behavioral conformity within a community of practice necessarily lead to homogeneity in the formation of lithic assemblages?
+To answer these questions, the quantitative analysis of process-level data associated with the product-level data becomes necessary.
+Behatrix (**Figure** \@ref(fig:behatrix), <https://www.boris.unito.it/behatrix/>), a sister software of BORIS, allows us to calculate the action sequence (dis)similarity between novice learners and expert demonstrators/fellow novice learners using established algorithms [see @cordoni2022 for an application of analyzing play behavior similarity among gorillas].
 
@@
 
+```{r behatrix, echo=FALSE, out.width = '150%', fig.align='center', fig.cap="The user interface of Behatrix displaying the transition probability between each action category in a handaxe knapping session."}
+knitr::include_graphics(here::here("figure/Fig3.jpg"))
+```
+
 ## Perception-level data
 
-Ethnographies revolving around experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S.
+Direct applications of ethnographic research in experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S.
 hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel.
-However, ethnography has never been formally recognized as a legitimate research method in experimental archaeology.
+However, ethnography has never been formally recognized as a legitimate research method in mainstream experimental archaeology.
 Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
@@
 Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two.
-As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project.
+As such, following data sharing principles of FAIR [@wilkinson2016; @nicholson2023] and CARE [@carroll2020; @gupta2023], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project.
 Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researchers; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
+
+On top of the digital data curation, an easily ignored but crucial aspect regarding the integrity and reliability of research in experimental archaeology is the long-term and proper curation of psychical specimens produced during experiments, which is particularly relevant to the product-level data.
+Haythron et al.
+[-@haythorn2018] sharply pointed out that re-running statistical analyses on a publicly available spreadsheet containing incorrect lithic projectile attribute data would be meaningless.
+In this case, a reexamination of the actual experiment samples becomes necessary.
+Moreover, it is likely that new research questions can only be answered through direct observation and measurement of the experimental assemblages themselves instead of data readily available in previously compiled spreadsheets [@eren2016].
+The existence of these possibilities thus requires that an experimental assemblage of interest should be curated in public institutions with easy access and detailed contextual information [@haythorn2018].
 
@@
 I thank Dietrich Stout for his helpful comments on earlier drafts of this article as well as Mark Moore and Shaya Jannati for inspiring discussions.
+I am also grateful to Metin Eren and two other anonymous reviewers for their insightful feedback.
+I would also like to express my sincere gratitude to Javier Baena Preysler, who helped proofread the Spanish translation of abstract and keywords.
+
+# Funding Statement
+
 This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."
 
-# References
+# Data Availability Statement
+
+No original data were used.
+
+# Competing Interests
+
+The author declare none.
+
+# Notes
+
+1.
+Strategy 2 is particularly relevant here, which is "the pursuit of general questions in present-day material culture in order to acquiring laws for making useful behavioral inferences" [@schiffer2010: 6]
+2.
+It should be noted that the definitions of internal and external validity constructs here solely concern the interpretation of experimental design and thus are different from the one previously proposed by Lycett and Eren [@eren2016; @lycett2013].
+
+# References Cited

### 2024-07-25 manuscript/manuscript_word.Rmd
title and readme file update.

@@
 Serving as a heuristic device, this alternative mode of knowledge production is highly flexible in nature, where each single component is detachable as dictated by individual research questions.
-\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"
+\\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production \\par \\ Abstracto: El objetivo de este artículo es ampliar el alcance de la arqueología experimental para centrar la atención en la variación multinivel y las interacciones entre los rangos de percepción, acciones y resultados.
+Este enfoque, formulado vagamente como el marco Percepción-Proceso-Producto (\"Triple P\"), ofrece una explicación más fundamentada y rica del registro arqueológico del pasado.
+Consta de tres principios 1) reconocer la compensación inherente entre control y generalizabilidad en el diseño de la investigación experimental; 2) fomentar proyectos de colaboración que impliquen a participantes en la investigación de procedencias distintas no limitados a los tradicionales participantes, como aficionados y novatos; 3) adoptar un flujo de trabajo que normalice la recogida y conservación de datos etológicos y etnográficos en proyectos experimentales.
+Como dispositivo heurístico, este modo alternativo de producción de conocimientos es muy flexible por naturaleza, y cada uno de sus componentes es desechable o modificable en función de las preguntas de investigación individuales.\\par \\ Palabras clave: Arqueología experimental; Análisis etológico; Investigación etnográfica; Maldición del conocimiento; Producción colaborativa de conocimientos"
 date: "`r Sys.Date()`"
@@
 bibliography: bibliography.bib
-csl: apa.csl
+csl: chicago-author-date.csl
 fontsize: 11pt
@@
 
-Built upon this critique of *Homo economicus* as well as the four strategies^[Strategy 2 is particularly relevant here, which is "the pursuit of general questions in present-day material culture in order to acquiring laws for making useful behavioral inferences" (Schiffer, 2010: 6)] of behavioral archaeology [@schiffer2010; @reid1975], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
+Built upon this critique of *Homo economicus* as well as the four strategies$^1$ of behavioral archaeology [@schiffer2010; @reid1975], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions.
 To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices.
@@
 
-The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267].
+The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@degtiar2023; @roe2009: 1266-1267]$^2$.
 Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022].
@@
 [-@stout2021] recorded the videos of expert flintknappers reproducing Oldowan and Acheulean technologies and then manually parsed their knapping activities using action grammar, generating multiple sequences of actions.
-Borrowing tools from computational linguistics, they then calculated the transition probability between each action category across two technological systems, which provided an objective and quantifiable proxy
-for measuring technological complexity.
+Borrowing tools from computational linguistics, they then calculated the transition probability between each action category across two technological systems, which provided an objective and quantifiable proxy for measuring technological complexity.
 Another scenario of its application is the measurement of behavioral similarity across individuals [@mobbs2021; @cristino2010], which is particularly relevant in the above-mentioned cultural transmission experiments.
@@
 
-```{r behatrix, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="The user interface of Behatrix displaying the transition probability between each action category in a handaxe knapping session."}
+```{r behatrix, echo=FALSE, out.width = '150%', fig.align='center', fig.cap="The user interface of Behatrix displaying the transition probability between each action category in a handaxe knapping session."}
 knitr::include_graphics(here::here("figure/Fig3.jpg"))
@@
 
-Ethnographies revolving around experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S.
+Direct applications of ethnographic research in experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S.
 hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel.
@@
 Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two.
-As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project.
+As such, following data sharing principles of FAIR [@wilkinson2016; @nicholson2023] and CARE [@carroll2020; @gupta2023], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project.
 Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researchers; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
@@
 Moreover, it is likely that new research questions can only be answered through direct observation and measurement of the experimental assemblages themselves instead of data readily available in previously compiled spreadsheets [@eren2016].
-The existence of these possibilities thus requires that an experimental assemblage of interest should be ideally curated in public institutions with easy access and detailed contextual information [@haythorn2018].
+The existence of these possibilities thus requires that an experimental assemblage of interest should be curated in public institutions with easy access and detailed contextual information [@haythorn2018].
 
@@
 I am also grateful to Metin Eren and two other anonymous reviewers for their insightful feedback.
+I would also like to express my sincere gratitude to Javier Baena Preysler, who helped proofread the Spanish translation of abstract and keywords.
+
+# Funding Statement
+
 This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."
 
-# References
+# Data Availability Statement
+
+No original data were used.
+
+# Competing Interests
+
+The author declare none.
+
+# Notes
+
+1.
+Strategy 2 is particularly relevant here, which is "the pursuit of general questions in present-day material culture in order to acquiring laws for making useful behavioral inferences" [@schiffer2010: 6]
+2.
+It should be noted that the definitions of internal and external validity constructs here solely concern the interpretation of experimental design and thus are different from the one previously proposed by Lycett and Eren [@eren2016; @lycett2013].
+
+# References Cited
